---
title: 'Assumptions and status'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assumptions and status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`assumptions()` returns choices recorded with a merchandising result.
`status_codes()` describes calculation diagnoses. `threads()` queries or changes the worker count for compiled equations. Currency labels use United States dollars (`USD`) in the example.

## Recorded settings

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load the example specifications
specifications <- example_products(name = 'douglas_fir')
```

```{r, echo = FALSE}
specifications %>%
  mutate(lengths = vapply(lengths, paste, character(1), collapse = ', ')) %>%
  select(product, lengths, min_length, max_length, min_sed, trim, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Select logs from the tree
result <- merchandise(dbh = tree$dbh,
                      ht = tree$ht,
                      model = tree$model,
                      species = tree$species,
                      id = tree$tree,
                      products = specifications,
                      currency = 'USD',
                      status = TRUE)
```

```{r}
## Extract the choices recorded with the result
recorded <- assumptions(x = result)
```

```{r, echo = FALSE}
recorded %>%
  select(assumption, species, model, value, units, source) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Records include equation mappings, defaults, units, and applicable species-property choices.
They omit local model coefficients. Preserve fitted objects or coefficient records with results using local equations.

## Calculation diagnoses

```{r}
## List the merchandising diagnoses
codes <- status_codes() %>%
  filter(source == 'merchandising')
```

```{r, echo = FALSE}
codes %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Count tree statuses in the result
checks <- result$trees %>%
  count(status)
```

```{r, echo = FALSE}
checks %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`status = TRUE` retains diagnoses and suppresses tree-status warnings.
Structural input errors still stop the call.
The `no_feasible_log` diagnosis denotes valid zero log volume. Other merchandising failures require review before reporting totals.

## Calculation workers

```{r}
## Query the current worker count
threads()

## Calculate diameters with a temporary worker count
with_threads(n = 4,
             code = dib(dbh = example_trees$dbh,
                        ht = example_trees$ht,
                        h = 20,
                        model = example_trees$model))
```

`with_threads()` restores the previous count when the expression finishes.
`MERCHANDISER_THREADS` sets the initial count when the package loads.
Equations written in R execute in the main R process.
