---
title: 'Assumptions and status'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assumptions and status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product order in `products()` sets cutting priority. Recorded assumptions and status codes support checking a run before summarizing its logs.

```{r}
## Select example trees
trees <- example_trees %>%
  slice_head(n = 3)

## Define domestic saw specifications
domestic_saw <- product(product = 'domestic_saw',  ## product name, unitless
                        min_length = 16,  ## feet
                        max_length = 40,  ## feet
                        trim = 1,  ## feet
                        min_sed = 6,  ## inches
                        volume_unit = 'scribner')  ## Scribner board feet

## Combine products in cutting priority order
specifications <- products(domestic_saw = domestic_saw)

## Cut and scale the trees
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = specifications)
```

## Status

Only trees with nonzero status appear in `status`. A clean run returns an empty table.

```{r}
## Show a clean status table
result$status

## Introduce a missing diameter
bad_trees <- trees %>%
  mutate(dbh = if_else(condition = tree_id == first(x = tree_id),
                       true = NA_real_,
                       false = dbh))

## Cut and scale the trees
bad_result <- merchandise(tree_id = bad_trees$tree_id,
                          dbh = bad_trees$dbh,
                          ht = bad_trees$ht,
                          spcd = bad_trees$spcd,
                          products = specifications)

## Show the failed input row
bad_result$status
```

Structural errors, such as duplicate identifiers or unknown product fields, stop the call.

```{r}
## Look up the status descriptions
codes <- status_codes()
```

```{r, echo = FALSE}
codes %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Assumptions

`assumptions()` returns equation assignments, stump heights, and any inferred bark ratios recorded for the trees.

```{r}
## Extract recorded assumptions
recorded <- assumptions(x = result)
```

```{r, echo = FALSE}
result$assumptions %>%
  mutate(source = if_else(assumption == 'species_model',
                          'pinned source library', source)) %>%
  rename(`value (unit column)` = value, `source (abbreviated)` = source) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Workers

```{r}
## Inspect the worker count
threads()

## Temporarily use a specified worker count
with_threads(n = 4,
             code = dib(dbh = trees$dbh,
                        ht = trees$ht,
                        h = 20,
                        spcd = trees$spcd)) %>%
  rename(`diameter inside bark (inches)` = value)
```

`with_threads()` restores the previous count. `MERCHANDISER_THREADS` sets the initial count when the package loads.
