---
title: 'Biomass, carbon, and green weight'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biomass, carbon, and green weight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`biomass()` estimates dry and green component masses, aboveground carbon, and carbon dioxide equivalent.
The National Scale Volume and Biomass system supplies these equations.
`green_weight()` converts an existing cubic volume using species properties.

## Biomass components

`nsvb_division_xy()` assigns ecological divisions from coordinates. The example inventory supplies longitude and latitude.

```{r}
## Assign ecological divisions to the shipped inventory
trees <- example_trees_pnw %>%
  mutate(division = nsvb_division_xy(x = longitude, y = latitude)) %>%
  slice_head(n = 5)
```

```{r, echo = FALSE}
trees %>%
  select(tree, species, dbh, ht_simulated, division) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Estimate component masses in pounds
mass <- biomass(dbh = trees$dbh,
                ht = trees$ht_simulated,
                spcd = trees$spcd,
                division = trees$division,
                id = trees$tree,
                status = TRUE)
```

```{r, echo = FALSE}
mass %>%
  select(id,
         dry_agb_no_foliage,
         dry_stem_wood,
         dry_stem_bark,
         dry_branches,
         dry_foliage,
         carbon,
         carbon_status) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`dry_` fields report oven-dry mass. `green_` fields include moisture.
Imperial calls return pounds, and metric calls return kilograms.
The aboveground total excludes foliage, and component fields overlap with totals.

The result includes stump, primary section, secondary section, tip, branch, and foliage components.
`decay_class` and `cull` apply the biomass system's tree-condition adjustments.
Its product limits are separate from merchandising specifications and selected cuts.

## One component and carbon dioxide equivalent

`biomass_component()` returns one named quantity per tree.
`co2e()` returns metric tons of carbon dioxide equivalent, regardless of the input measurement system.

```{r}
## Add dry stem mass and carbon dioxide equivalent to tree records
carbon <- trees %>%
  transmute(tree,
            dry_wood_lb = biomass_component(dbh = dbh,
                                            ht = ht_simulated,
                                            spcd = spcd,
                                            division = division,
                                            component = 'dry_stem_wood'),
            carbon_dioxide_t = co2e(dbh = dbh,
                                    ht = ht_simulated,
                                    spcd = spcd,
                                    division = division),
            fraction = carbon_fraction(spcd = spcd))
```

```{r, echo = FALSE}
carbon %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Carbon excludes foliage. `carbon_fraction()` returns the live-tree species fraction with a source attribute.
Dead-tree carbon follows the decay-specific rules in `biomass()`.
The `co2e` column in `biomass()` uses pounds or kilograms, unlike the separate `co2e()` function.

## Weight from cubic volume

`green_weight()` uses species wood density, bark volume, and moisture properties.
Its default returns green wood and attached bark weight from inside-bark cubic volume.

```{r}
## Convert modeled stem volume to green pounds
weight <- example_trees %>%
  mutate(cubic = stem_volume(dbh = dbh, ht = ht, model = model),
         green_lb = green_weight(volume = cubic, spcd = spcd),
         dry_wood_lb = green_weight(volume = cubic,
                                    spcd = spcd,
                                    component = 'wood',
                                    moisture = 'dry')) %>%
  select(tree, species, cubic, green_lb, dry_wood_lb)
```

```{r, echo = FALSE}
weight %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Property overrides include `specific_gravity`, `moisture_pct`, and corresponding bark fields.
`volume_basis = 'outside'` accepts volume including bark.
This volume conversion is independent of the fitted biomass system.
