---
title: 'Biomass and carbon'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biomass and carbon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

`biomass()` returns dry biomass, carbon, and carbon dioxide equivalent in metric tonnes from inputs in inches and feet.
Component columns support tree and stand mass summaries.

## National coefficients

The default uses national coefficients. Each output row corresponds to its input tree.

```{r}
## Calculate biomass for the shipped tree list
mass <- example_trees_pnw %>%
  mutate(biomass(dbh = dbh,
                 ht = ht,
                 spcd = spcd))
```

```{r, echo = FALSE}
mass %>%
  slice_head(n = 6) %>%
  transmute(tree_id = tree_id, `stem wood (tonnes)` = dry_stem_wood,
            `stem bark (tonnes)` = dry_stem_bark, `branches (tonnes)` = dry_branches,
            `foliage (tonnes)` = dry_foliage, `carbon (tonnes)` = carbon,
            `carbon dioxide equivalent (tonnes)` = tco2e, status) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The component curves summarize mass by diameter class for each species.

```{r, echo = FALSE, fig.width = 9}
components <- mass %>%
  mutate(diameter_class = floor(dbh / 2) * 2) %>%
  group_by(spcd, diameter_class) %>%
  summarize(across(c(dry_stem_wood, dry_stem_bark, dry_branches, dry_foliage), mean),
            .groups = 'drop')
components %>%
  rename(`diameter class lower bound (inches)` = diameter_class,
         `mean stem wood (tonnes)` = dry_stem_wood,
         `mean stem bark (tonnes)` = dry_stem_bark,
         `mean branches (tonnes)` = dry_branches,
         `mean foliage (tonnes)` = dry_foliage) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE, fig.width = 9}
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
colors <- c('#0072B2', '#E69F00', '#009E73', '#CC79A7')
for (code in unique(components$spcd)) {
  z <- components %>% filter(spcd == code)
  matplot(z$diameter_class, z[3:6], type = 'l', lty = 1, lwd = 2, col = colors,
          xlab = 'Diameter class lower bound (inches)', ylab = 'Mean dry mass (tonnes)')
  legend('topleft', legend = c(paste('Species', code), 'Stem wood', 'Stem bark',
                               'Branches', 'Foliage'),
         col = c(NA, colors), lty = c(NA, rep(1, 4)), bty = 'n', cex = 0.8)
}
```

`dry_aboveground_no_foliage` excludes foliage. Stem totals overlap stump, saw, topwood, and tip columns, so those columns are not additive.
Carbon excludes foliage. `tco2e` converts carbon to carbon dioxide equivalent.

Invalid inputs retain their rows with missing masses and a status code.

## Carbon fraction

`carbon_fraction()` returns the live tree carbon fraction of dry mass, with a source attribute.

```{r}
## Look up the example species fractions
fractions <- example_trees %>%
  distinct(spcd) %>%
  mutate(carbon_fraction = carbon_fraction(spcd = spcd))
```

```{r, echo = FALSE}
fractions %>%
  rename(`carbon fraction (unitless)` = carbon_fraction) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Optional location lookup

[`nsvb_division()`](../reference/nsvb_division.html) returns county ecological codes. [`nsvb_division_xy()`](../reference/nsvb_division_xy.html) returns coordinate assignments and status, using the shipped boundaries.
Use county codes directly, or the coordinate result's `value` column, as the optional coefficient selector.
