---
title: 'Bucking choices and prices'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bucking choices and prices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`merchandise()` selects the longest eligible log within the reachable product sequence.
`optimize_bucking()` selects cuts for value or net inside-bark cubic volume under the same specifications.

## Optimize a tree

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load the example specifications
specifications <- example_products(name = 'douglas_fir')
```

```{r, echo = FALSE}
specifications %>%
  mutate(lengths = vapply(lengths, paste, character(1), collapse = ', ')) %>%
  select(product, lengths, min_length, max_length, min_sed, trim, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`price` is currency per `price_per` measurement units. The shipped prices are illustrative and use United States dollars (`USD`).

```{r, echo = FALSE}
specifications %>%
  select(product, volume_unit, price, price_per) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Select logs from the tree
result <- optimize_bucking(dbh = tree$dbh,
                           ht = tree$ht,
                           model = tree$model,
                           species = tree$species,
                           id = tree$tree,
                           products = specifications,
                           currency = 'USD',
                           status = TRUE,
                           objective = 'value')
```

```{r, echo = FALSE}
result$logs %>%
  select(log, product, nominal_length, sed_ib, net_scale, scale_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Draw the optimized cuts
plot(result)
```

`objective = 'auto'` selects value when prices are present and net cubic volume otherwise.
For value optimization, `unpriced = 'zero'` assigns zero value to unpriced products.
`unpriced = 'error'` requires prices for every product.

## Compare price scenarios

`compare_bucking_prices()` values existing cuts at each scenario's prices, then optimizes the trees again.
Every scenario must contain every product exactly once.

```{r}
## Increase the example pulp price while retaining sawlog prices
scenarios <- specifications %>%
  transmute(scenario = 'higher_pulp',
            product,
            price = case_when(volume_unit == 'green_ton' ~ price * 2,
                              .default = price))
```

```{r, echo = FALSE}
scenarios %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Compare fixed cuts with cuts selected for the new prices
comparison <- compare_bucking_prices(x = result, price_scenarios = scenarios)
```

```{r, echo = FALSE}
comparison %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`repriced_value` values the original cuts. `rebucked_value` values the newly selected cuts.
`rebuck_gain` is their difference. Comparisons use unexpanded totals in each scenario's currency.

The `rebucked_results` attribute retains each new result.
Compare prices before applying later percentage deductions. Local models must remain registered for the repeated calculations.
