---
title: 'Bucking strategy and value'
output:
  rmarkdown::html_vignette:
    md_extensions: -smart
vignette: >
  %\VignetteIndexEntry{Bucking strategy and value}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      dev.args = list(bg = 'white'), fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Bucking strategy selects logs by product priority or by their combined value. Compare the strategies when accepted lengths and prices offer competing cuts within the same stem. This calculation uses the Get started tree and its stored taper equation, with dimensions in inches and feet and illustrative prices in United States dollars.

## What prices and dimensions compete?

The cascade offers each available section to products in the order supplied to `products()`. It takes the longest feasible sawlog at each step before assigning remaining eligible wood to pulp. Prices value those selected logs without changing their priority. Optimization compares combinations of eligible cuts within each available segment.

```{r}
## Attach the calculation and data tools
library(merchandiser)
library(dplyr)
```

```{r}
## Select the tree with a stored taper equation
tree <- example_trees %>%
  filter(tree_id == 5)
```


```{r}
## Accept sawlogs across the nominal length range
sawlog <- product(product = 'Sawlog',  ## product label
                  min_length = 16,  ## feet
                  max_length = 40,  ## feet
                  min_sed = 6,  ## inches
                  inside_bark = TRUE,  ## limits measured inside bark
                  trim = 0.5,  ## feet
                  volume_unit = 'scribner',  ## board feet
                  price = 600,  ## dollars
                  price_per = 1000)  ## board feet

## Assign remaining eligible sections to pulp by green weight
pulp <- product(product = 'Pulp',  ## product label
                min_length = 8,  ## feet
                max_length = 20,  ## feet
                min_sed = 3,  ## inches
                inside_bark = TRUE,  ## limits measured inside bark
                trim = 0.5,  ## feet
                volume_unit = 'green_ton',  ## green short tons
                price = 30,  ## dollars
                price_per = 1)  ## green short tons
```


```{r}
## Set the cascade's product priority
specifications <- products(sawlog, pulp)
```


```{r}
## Select logs in the supplied product order
cascade <- merchandise(tree_id = tree$tree_id,
                       dbh = tree$dbh,
                       ht = tree$ht,
                       spcd = tree$spcd,
                       products = specifications,
                       model = tree$model,
                       strategy = 'cascade')
```


```{r}
## Select the combination with the greatest value
optimized <- merchandise(tree_id = tree$tree_id,
                         dbh = tree$dbh,
                         ht = tree$ht,
                         spcd = tree$spcd,
                         products = specifications,
                         model = tree$model,
                         strategy = 'optimize')
```

## Which strategy assigns the greater value?

Every product must have a finite positive price for optimization. `price_per` puts that price on the product's scale basis, allowing board foot and weight products to contribute to one dollar objective. The scale itself remains in each product's own units.


```{r}
## Compare dollar totals without combining unlike scale units
values <- bind_rows(cascade = cascade$logs,
                    optimized = optimized$logs,
                    .id = 'strategy') %>%
  group_by(strategy) %>%
  summarize(logs = n(),
            board_feet = sum(scale[volume_unit == 'scribner']),
            value = sum(value))
```


```{r, echo = FALSE}
values %>%
  rename(`logs (count)` = logs, `scale (Scribner board feet)` = board_feet,
         `value (dollars)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```


Cascade selects `r values$logs[values$strategy == 'cascade']` logs totaling `r values$board_feet[values$strategy == 'cascade']` Scribner board feet and about `r round(values$value[values$strategy == 'cascade'])` dollars, while optimization selects `r values$logs[values$strategy == 'optimized']` logs totaling `r values$board_feet[values$strategy == 'optimized']` board feet and about `r round(values$value[values$strategy == 'optimized'])` dollars. Log counts and dollar totals include pulp, while the board foot totals include only sawlogs.

## Where did the cuts change?

Each candidate occupies nominal length plus trim, while scale uses the nominal body. Under the small end diameter rule, a shorter log carries a larger small end diameter on this tapering stem. The longest feasible log therefore need not produce the highest total scale. Here, shorter sawlogs increase the board foot total despite the additional trim allowances. The drawings retain the same colors for each product.


```{r, eval = FALSE}
## Compare the selected cuts on the same stem
par(mfrow = c(1, 2))
plot(cascade)
plot(optimized)
```

```{r, echo = FALSE, fig.width = 12}
par(mfrow = c(1, 2))
plot(cascade, main = '', ylab = 'Stem radius (inches)', yaxt = 's')
plot(optimized, main = '', ylab = 'Stem radius (inches)', yaxt = 's')
```

The cascade drawing starts with a `r cascade$logs$length[1]`-foot sawlog, while the optimized drawing starts with a `r optimized$logs$length[1]`-foot sawlog and contains `r nrow(optimized$logs)` selected logs.


```{r, echo = FALSE}
bind_rows(cascade = cascade$logs, optimized = optimized$logs, .id = 'strategy') %>%
  transmute(strategy, log, product, `start (feet)` = start_height,
            `end (feet)` = end_height, `nominal length (feet)` = length,
            `small-end diameter (inches)` = sed,
            `scale (Scribner board feet)` = if_else(volume_unit == 'scribner', scale, NA_real_),
            `scale (green short tons)` = if_else(volume_unit == 'green_ton', scale, NA_real_),
            `value (dollars)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The first optimized sawlog has a physical small end diameter of `r round(optimized$logs$sed[1], 2)` inches, compared with `r round(cascade$logs$sed[1], 2)` inches for the first cascade sawlog.

## How are equal values resolved?

```{r, echo = FALSE}
length_grid <- formals(stem_profile)$step
merchandise_rd <- tools::Rd_db('merchandiser')[['merchandise.Rd']]
merchandise_help <- paste(capture.output(tools::Rd2txt(merchandise_rd)), collapse = ' ')
merchandise_help <- gsub('[[:space:]]+', ' ', merchandise_help)
tie_pattern <- 'within [a-z]+ machine epsilons times the larger absolute value'
tie_rule <- regmatches(merchandise_help, regexpr(tie_pattern, merchandise_help))
position_match <- regexpr('[0-9,]+(?= cut positions)', merchandise_help, perl = TRUE)
position_limit <- regmatches(merchandise_help, position_match)
```

The optimizer uses dynamic programming over reachable cut positions and eligible nominal lengths on a `r length_grid`-foot grid. As documented in `merchandise()`, values `r tie_rule` are treated as ties. Among tied values it favors fewer logs, then compares nominal lengths from the lower stem upward, favoring the first longer log. Remaining ties favor the first lower physical end in the same order. The final comparison uses byte order of serialized start heights, physical ends, cut kinds, nominal lengths, and product names along each path. These comparisons do not round the returned dollar values.

## What limits computation?

The state includes cut position and the counts of products with `max_logs` limits. Cull and restriction boundaries divide the available stem into segments and restart those counts. The algorithm compares eligible continuations within that structure, rather than moving defects or changing the supplied stem equation.

Without count limits, work grows with positions, products, and candidate lengths. Retaining the chosen paths also costs time and memory proportional to their log counts. Independently capped products multiply the possible states by their count combinations, making the worst case exponential in the number of capped products. Only reached states are stored.

Preparation rejects an off-grid cut-position expansion exceeding `r position_limit` cut positions, the limit documented in `merchandise()`. Broad length ranges and several independent log caps can therefore make a specification expensive even for a modest tree list. Trees use the package's parallel engine, but additional threads do not reduce the number of candidate states required for a tree.
