---
title: 'Compiling an inventory'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compiling an inventory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`stand_table()` expands tree counts, basal area, log counts, and product quantities.
`stock_table()` uses the same calculation with product and diameter-class grouping by default.
Both require the complete inventory and supplied expansion factors.

```{r}
## Use the complete simulated heights and retain tree identifiers
trees <- example_trees_pnw %>%
  mutate(id = tree, ht = ht_simulated)
```

```{r}
## Load the inventory specifications
specifications <- example_products(name = 'pnw')
```

```{r}
## Apply the product table to the inventory
result <- merchandise(dbh = trees$dbh,
                      ht = trees$ht,
                      species = trees$species,
                      id = trees$id,
                      products = specifications,
                      status = TRUE)
```

## Stand and stock tables

```{r}
## Expand tree counts and quantities by species
stand <- stand_table(x = result,
                     trees = trees,
                     expansion = trees$expansion_factor,
                     by = 'species')
```

```{r, echo = FALSE}
stand %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Expand stock by product
stock <- stock_table(x = result,
                     trees = trees,
                     expansion = trees$expansion_factor,
                     by = 'product')
```

```{r, echo = FALSE}
stock %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`trees$id` joins the inventory to the result. Expansion follows the caller's inventory row order.
`area_unit` labels expansion without converting it. These are weighted sums without variance estimates.

A tree contributing to multiple products appears in each product group.
Omit `product` from grouping to count each tree once in its inventory group.
Different measurement units receive separate quantity columns.

## Product summaries

`product_summary()` summarizes a saved result. `table = 'logs'` groups by product and measurement fields.
`table = 'trees'` summarizes tree quantities. `na_action = 'propagate'` retains missing quantities in totals.

```{r}
## Sum log quantities by product
summary <- product_summary(x = result,
                           table = 'logs',
                           expansion = trees$expansion_factor,
                           na_action = 'propagate')
```

```{r, echo = FALSE}
summary %>%
  select(product, scale_unit, input_trees, net_scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Calculate product totals directly from inventory measurements
direct <- product_summary_by_tree(dbh = trees$dbh,
                                  ht = trees$ht,
                                  species = trees$species,
                                  products = specifications,
                                  status = TRUE)
```

```{r, echo = FALSE}
direct %>%
  select(product, scale_unit, input_trees, net_scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The direct call uses unexpanded totals unless `expansion` is supplied.
`merch_summary()` is the deprecated name for `product_summary()` and returns the same table.

```{r}
## Summarize a saved result through the compatibility name
legacy <- merch_summary(x = result, table = 'trees')
```

```{r, echo = FALSE}
legacy %>%
  select(input_trees, log_count, log_net_cubic_ib) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Add selected log counts to the complete example tree list
counts <- example_trees %>%
  transmute(tree,
            logs = tree_log_count(dbh = dbh,
                                  ht = ht,
                                  species = species,
                                  model = model,
                                  products = specifications))
```

```{r, echo = FALSE}
counts %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Stand and stock tables retain valid zero-log trees, failures, and trees absent from the result.
`merchandising_status` separates those groups. Missing quantities propagate within applicable groups.
