---
title: 'Recording defect'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recording defect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`defect()` supplies located restrictions that can change log selection. `apply_defect_pct()` deducts percentages after bucking and retains the cuts.

The example currency is United States dollars (`USD`).

## Located conditions

`defect()` records a condition by tree identifier and height. Heights use the same units as the merchandising call.

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Record rot along the example stem
record <- defect(id = tree$tree,
                 from = 10,
                 to = 30,
                 effect = 'rot',
                 percent = 20)
```

```{r, echo = FALSE}
record %>%
  select(id, from, to, effect, percent) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load specifications and validate the record against the tree
specifications <- example_products(name = 'douglas_fir')

## Check defect bounds and fields
checked <- validate_defects(x = record,
                            products = specifications,
                            ht = tree$ht,
                            id = tree$tree)
```

```{r, echo = FALSE}
checked %>%
  select(id, from, to, effect, percent, status) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Select logs from the tree
result <- merchandise(dbh = tree$dbh,
                      ht = tree$ht,
                      model = tree$model,
                      species = tree$species,
                      id = tree$tree,
                      products = specifications,
                      currency = 'USD',
                      status = TRUE,
                      defects = record)
```

```{r, echo = FALSE}
result$logs %>%
  select(log, product, nominal_length, sed_ib, net_scale, scale_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Draw the cuts and located defect
plot(result)
```

Rot deductions use overlapping inside-bark volume. Overlapping rot percentages use the greatest percentage.
Cull intervals remove stem sections and restart product selection above them.
Pulp intervals restrict eligibility to the product with `accepts_pulp_restriction = TRUE`.

Sweep and crook use caller-defined ordered categories. They restrict eligibility without assigning a percentage deduction.
A break ends selection at its height. A fork applies a pulp restriction above its height.

## Stopping heights

`defects_from_stoppers()` converts recorded stopping heights into cull and pulp intervals.
A jump butt removes the lower interval. A saw stop starts a pulp restriction, and a pulp stop removes the remaining top.

```{r}
## Select complete example measurements with stopping heights
stopped <- example_trees_south %>%
  filter(!is.na(saw_stop) | !is.na(pulp_stop) | !is.na(jump_butt)) %>%
  slice_head(n = 3)
```

```{r, echo = FALSE}
stopped %>%
  select(tree, ht_simulated, saw_stop, pulp_stop, jump_butt) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Convert stopping heights to located records
intervals <- defects_from_stoppers(id = stopped$tree,
                                   ht = stopped$ht_simulated,
                                   saw_stop = stopped$saw_stop,
                                   pulp_stop = stopped$pulp_stop,
                                   jump_butt = stopped$jump_butt)
```

```{r, echo = FALSE}
intervals %>%
  select(id, from, to, effect) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Percentages

`defect_pct_from_thirds()` weights supplied percentages by modeled inside-bark volume in each third of total height.


```{r}
## Combine example percentages by stem-third volume
percentage <- defect_pct_from_thirds(dbh = tree$dbh,
                                     ht = tree$ht,
                                     model = tree$model,
                                     pct_lower = 10,
                                     pct_middle = 20,
                                     pct_upper = 30,
                                     status = TRUE)
```

```{r, echo = FALSE}
percentage %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Apply the percentage to current net quantities
deducted <- apply_defect_pct(x = result,
                             pct = percentage$value,
                             kind = 'thirds')
```

```{r, echo = FALSE}
deducted$logs %>%
  select(log, gross_scale, located_net_scale, net_scale, scale_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`apply_defect_pct()` returns an updated result. It updates physical quantities, measurements, values, and deduction records.
Repeated calls compound on current net quantities.

Missing quantities remain missing. Missing or invalid percentages stop the call.
