---
title: 'Recording defect'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recording defect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product order in `products()` sets cutting priority. Located defect records change available stem sections and product eligibility.

```{r}
## Define export specifications
export <- product(product = 'export',  ## product name, unitless
                  spcd = c(202, 263),  ## species codes, unitless
                  min_dbh = 12,  ## inches
                  min_length = 32,  ## feet
                  max_length = 40,  ## feet
                  trim = 1,  ## feet
                  min_sed = 12,  ## inches
                  inside_bark = TRUE,  ## logical, unitless
                  volume_unit = 'scribner',  ## board feet
                  split_scale = FALSE)  ## logical, unitless

## Define domestic saw specifications
domestic_saw <- product(product = 'domestic_saw',  ## product name, unitless
                        spcd = c(202, 263),  ## species codes, unitless
                        min_dbh = 10,  ## inches
                        min_length = 16,  ## feet
                        max_length = 40,  ## feet
                        trim = 1,  ## feet
                        min_sed = 6,  ## inches
                        inside_bark = TRUE,  ## logical, unitless
                        volume_unit = 'scribner',  ## board feet
                        split_scale = FALSE)  ## logical, unitless

## Define pulp specifications
pulp <- product(product = 'pulp',  ## product name, unitless
                spcd = c(202, 263),  ## species codes, unitless
                min_dbh = 5,  ## inches
                min_length = 8,  ## feet
                max_length = 40,  ## feet
                trim = 0.5,  ## feet
                min_sed = 3,  ## inches
                inside_bark = TRUE,  ## logical, unitless
                volume_unit = 'green_ton')  ## green short tons

## Combine products in cutting priority order
specifications <- products(export = export,
                           domestic_saw = domestic_saw,
                           pulp = pulp)
```

## Effects

`cull` removes an interval. `restrict` allows only its named product within an interval. Both split the stem and restart priority.

```{r}
## Record a cull on an example tree
defect(tree_id = example_trees_pnw$tree_id[7],
       start_height = 1,
       end_height = 8,
       effect = 'cull') %>%
  rename(`start_height (feet)` = start_height,
         `end_height (feet)` = end_height,
         `percent (percent)` = percent)

## Record a product restriction
defect(tree_id = example_trees_pnw$tree_id[28],
       start_height = 35,
       end_height = NA,
       effect = 'restrict',
       product = 'pulp') %>%
  rename(`start_height (feet)` = start_height,
         `end_height (feet)` = end_height,
         `percent (percent)` = percent)
```

`end` stops cutting at `start_height`. `sweep` rejects overlapping logs only when its percentage exceeds the product's `max_sweep`.

```{r}
## Record an end
defect(tree_id = example_trees_pnw$tree_id[21],
       start_height = 88,
       end_height = NA,
       effect = 'end') %>%
  rename(`start_height (feet)` = start_height,
         `end_height (feet)` = end_height,
         `percent (percent)` = percent)

## Record sweep severity
defect(tree_id = example_trees_pnw$tree_id[35],
       start_height = 28,
       end_height = 70,
       effect = 'sweep',
       percent = 25) %>%
  rename(`start_height (feet)` = start_height,
         `end_height (feet)` = end_height,
         `percent (percent)` = percent)
```

Missing `end_height` extends the record to the tree top. No defect percentage is applied within `merchandise()`.

## Mapped defects

```{r}
## Check the shipped defect records
checked <- validate_defects(
  defects = example_defects_pnw,
  products = specifications,
  ht = example_trees_pnw$ht,
  tree_id = example_trees_pnw$tree_id
)
```

```{r, echo = FALSE}
checked %>%
  slice_head(n = 6) %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height,
         `percent (%)` = percent) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Select the tree with the first shipped cull
trees <- example_trees_pnw %>%
  filter(tree_id == first(x = example_defects_pnw$tree_id))

## Select its mapped defects
records <- example_defects_pnw %>%
  filter(tree_id %in% trees$tree_id)

## Cut and scale the trees
clean <- merchandise(tree_id = trees$tree_id,
                     dbh = trees$dbh,
                     ht = trees$ht,
                     spcd = trees$spcd,
                     products = specifications)

## Cut and scale the trees
affected <- merchandise(tree_id = trees$tree_id,
                        dbh = trees$dbh,
                        ht = trees$ht,
                        spcd = trees$spcd,
                        products = specifications,
                        defects = records)
```

```{r, echo = FALSE}
bind_rows(clean = clean$logs, affected = affected$logs, .id = 'run') %>%
  filter(volume_unit == 'scribner') %>%
  transmute(run, tree_id, log, product,
            `start height (feet)` = start_height, `end height (feet)` = end_height,
            `length (feet)` = length, `small end diameter (inches)` = sed,
            `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
bind_rows(clean = clean$logs, affected = affected$logs, .id = 'run') %>%
  filter(volume_unit == 'green_ton') %>%
  transmute(run, tree_id, log, product,
            `start height (feet)` = start_height, `end height (feet)` = end_height,
            `length (feet)` = length, `small end diameter (inches)` = sed,
            `scale (green short tons)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

<details>
<summary>Drawing measurements</summary>

```{r, echo = FALSE}
drawing_trees <- trees
drawing_rows <- match(drawing_trees$tree_id, clean$call$tree_id)
drawing_models <- clean$call$model[drawing_rows]
drawing_profiles <- stem_profile(
  tree_id = drawing_trees$tree_id,
  dbh = drawing_trees$dbh,
  ht = drawing_trees$ht,
  spcd = drawing_trees$spcd,
  model = drawing_models,
  step = 0.5,
  from = 0
)
drawing_profiles %>%
  transmute(tree_id, `height (feet)` = h, `radius inside bark (inches)` = dib / 2) %>%
  knitr::kable(digits = 4, row.names = FALSE)
bind_rows(clean = clean$residuals, affected = affected$residuals, .id = 'run') %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height) %>%
  knitr::kable(digits = 3, row.names = FALSE)
records %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height,
         `percent (percent)` = percent) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

</details>

The cull moves the first usable cut upward. The left drawing omits defects, and the right drawing uses the shipped record.

```{r, echo = FALSE, fig.width = 12, fig.height = 4.5}
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
plot(clean, main = '', ylab = 'Stem radius (inches)')
plot(affected, main = '', ylab = 'Stem radius (inches)')
```

## Stopping heights

`defects_from_stoppers()` converts jump butts to culls, saw stops to restrictions, and pulp stops to ends.
The [southern workflow](southern.html) applies the converted rows to the complete list.

```{r}
## Convert the shipped stopper columns
stoppers <- defects_from_stoppers(
  tree_id = example_trees_south$tree_id,
  ht = example_trees_south$ht,
  topwood_product = "pulpwood",
  saw_stop = example_trees_south$saw_stop,
  pulp_stop = example_trees_south$pulp_stop,
  jump_butt = example_trees_south$jump_butt
)
```

```{r, echo = FALSE}
stoppers %>%
  slice_head(n = 5) %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height,
         `percent (%)` = percent) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Stem thirds

`defect_by_thirds()` weights percentages by modeled inside bark volume in each third of total height.
It returns a standalone percentage and status. Applying the percentage is a separate script calculation.

```{r}
## Combine percentages using species default taper
defect_by_thirds(dbh = trees$dbh,
                 ht = trees$ht,
                 spcd = trees$spcd,
                 lower = 10,
                 middle = 20,
                 upper = 0) %>%
  rename(`defect (percent)` = value)
```
