---
title: 'Defining products'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining products}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`product()` creates a specification row. `products()` combines rows and validates the resulting table.
Lower `priority` values are considered first.

## Product fields

```{r}
## Define a sawlog specification
saw <- product(product = 'saw',
               priority = 1,
               lengths = c(16, 32),
               trim = 0.5,
               min_sed = 6,
               volume_unit = 'scribner',
               inside_bark = TRUE)
```

```{r, echo = FALSE}
saw %>%
  select(product, priority, lengths, trim, min_sed, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Define a pulp specification
pulp <- product(product = 'pulp',
                priority = 2,
                min_length = 8,
                max_length = 20,
                length_step = 1,
                trim = 0.5,
                min_sed = 3,
                volume_unit = 'green_ton',
                accepts_pulp_restriction = TRUE)
```

```{r}
## Combine and check the specifications
specifications <- products(saw, pulp) %>%
  validate_products()
```

```{r, echo = FALSE}
specifications %>%
  select(product, priority, min_sed, volume_unit, accepts_pulp_restriction) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Use `lengths` for discrete lengths or `min_length`, `max_length`, and `length_step` for a range.
`min_sed` tests the physical small end, including trim. `inside_bark` chooses the diameter basis for acceptance.

Tree entry checks species, diameter, age, and pruning restrictions.
`allow_after_higher_product` controls later use after an earlier product.
`allow_lower_products` controls progression to subsequent products.
[Product fields](../reference/product_schema.html) control tree entry, log acceptance, and measurement.

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Apply the specifications to the example tree
result <- merchandise(dbh = tree$dbh,
                      ht = tree$ht,
                      model = tree$model,
                      species = tree$species,
                      products = specifications,
                      status = TRUE)
```

```{r, echo = FALSE}
result$logs %>%
  select(log, product, nominal_length, sed_ib, net_scale, scale_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Shipped specifications

`example_product_names()` lists the available tables. `example_products()` loads one.
`product_presets()` and `product_preset()` are compatibility names for the same operations.

```{r}
## List the example tables
available <- example_product_names()
```

```{r, echo = FALSE}
available %>%
  select(name, status) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load the example specifications
specifications <- example_products(name = 'pnw')
```

```{r, echo = FALSE}
specifications %>%
  mutate(lengths = vapply(lengths, paste, character(1), collapse = ', ')) %>%
  select(product, lengths, min_length, max_length, min_sed, trim, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load the same specifications through the compatibility names
preset_names <- product_presets()

## Load a preset table
preset_products <- product_preset(name = 'pnw')
```

```{r, echo = FALSE}
preset_products %>%
  select(product, priority, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Source rule records

`nvel_rules()` records National Volume Estimator Library settings.
`products_from_nvel_rules()` translates a complete record into products and accompanying stem limits.


```{r}
## Record the source segmentation and measurement settings
rules <- nvel_rules(even_or_odd = 2,
                    option = 14,
                    maximum_length = 40,
                    minimum_length = 8,
                    minimum_top_length = 8,
                    merchantable_length = 80,
                    primary_top = 6,
                    secondary_top = 4,
                    stump = 1,
                    trim = 0.5,
                    bark_ratio = 0.9,
                    minimum_board_foot_dbh = 1,
                    scribner = 'table')
```

```{r}
## Translate the complete rule record
translated <- products_from_nvel_rules(x = rules)
```

```{r, echo = FALSE}
translated$products %>%
  select(product, min_sed, min_length, max_length, scale_rule) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Translated products retain expanded source measurement fields.
The returned list also contains `stump_ht`, `utilization_height`, and `model_aux`.
The translated height cap adds `stump` to `merchantable_length`.
Both translated products retain board-foot measurement, including the product accepting pulp restrictions.
