---
title: 'Defining products'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining products}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product order in `products()` sets cutting priority. Product fields set the trees and log sections eligible for each output.

```{r}
## Select example trees
trees <- example_trees %>%
  slice_head(n = 3)

## Define export specifications
export <- product(product = 'export',  ## product name, unitless
                  spcd = c(202, 263),  ## species codes, unitless
                  min_dbh = 12,  ## inches
                  min_length = 32,  ## feet
                  max_length = 40,  ## feet
                  trim = 1,  ## feet
                  min_sed = 12,  ## inches
                  inside_bark = TRUE,  ## logical, unitless
                  volume_unit = 'scribner',  ## board feet
                  split_scale = FALSE)  ## logical, unitless

## Define domestic saw specifications
domestic_saw <- product(product = 'domestic_saw',  ## product name, unitless
                        spcd = c(202, 263),  ## species codes, unitless
                        min_dbh = 10,  ## inches
                        min_length = 16,  ## feet
                        max_length = 40,  ## feet
                        trim = 1,  ## feet
                        min_sed = 6,  ## inches
                        inside_bark = TRUE,  ## logical, unitless
                        volume_unit = 'scribner',  ## board feet
                        split_scale = FALSE)  ## logical, unitless

## Define pulp specifications
pulp <- product(product = 'pulp',  ## product name, unitless
                spcd = c(202, 263),  ## species codes, unitless
                min_dbh = 5,  ## inches
                min_length = 8,  ## feet
                max_length = 40,  ## feet
                trim = 0.5,  ## feet
                min_sed = 3,  ## inches
                inside_bark = TRUE,  ## logical, unitless
                volume_unit = 'green_ton')  ## green short tons

## Combine products in cutting priority order
specifications <- products(export = export,
                           domestic_saw = domestic_saw,
                           pulp = pulp)

## Cut and scale the trees
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = specifications,
                      model = trees$model)
```

```{r, echo = FALSE}
result$logs %>%
  filter(volume_unit == 'scribner') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `length (feet)` = length,
            `sed (inches)` = sed,
            `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
result$logs %>%
  filter(volume_unit == 'green_ton') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `length (feet)` = length,
            `sed (inches)` = sed,
            `scale (green short tons)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Tree and log limits

`min_dbh` and `max_dbh` apply to tree diameter. `min_age` and `max_age` apply to supplied tree age.
Lower limits are inclusive. Upper tree diameter and age limits are exclusive.

Missing age prevents eligibility for products with age limits.

`min_length` and `max_length` bound candidate nominal lengths. Candidates use the internal half foot grid.
`min_sed`, `max_sed`, `min_led`, and `max_led` constrain physical end diameters.
`max_logs` limits product counts within each segment.

`requires_pruned` requires the entire cut below `pruned_ht`. `max_sweep` rejects overlaps above its percentage limit.

```{r}
## Define clear specifications
clear <- product(product = 'clear',  ## product name, unitless
                 min_age = 35,  ## years
                 max_age = 70,  ## years
                 requires_pruned = TRUE,  ## logical, unitless
                 min_length = 16,  ## feet
                 max_length = 16,  ## feet
                 min_sed = 6,  ## inches
                 max_sweep = 10,  ## percent
                 volume_unit = 'cubic')  ## cubic feet

## Cut and scale the trees
clear_result <- merchandise(tree_id = trees$tree_id,
                            dbh = trees$dbh,
                            ht = trees$ht,
                            spcd = trees$spcd,
                            products = clear,
                            model = trees$model,
                            age = 45,
                            pruned_ht = 34)
```

```{r, echo = FALSE}
clear_result$logs %>%
  filter(volume_unit == 'cubic') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `length (feet)` = length,
            `sed (inches)` = sed,
            `scale (cubic feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`age` and `pruned_ht` supply scenario measurements.
[Southern stands](southern.html) compares explicit age limits across the shipped stand ages.

## Cords

`cord_solid_fraction` supplies the solid wood proportion for the stacked cord conversion.

```{r}
## Define cordwood specifications
cordwood <- product(product = 'cordwood',  ## product name, unitless
                    min_length = 8,  ## feet
                    max_length = 16,  ## feet
                    min_sed = 3,  ## inches
                    inside_bark = TRUE,  ## logical, unitless
                    volume_unit = 'cord',  ## cords
                    cord_solid_fraction = 0.7)  ## solid wood proportion, unitless

## Cut and scale the trees
cord_result <- merchandise(tree_id = trees$tree_id,
                           dbh = trees$dbh,
                           ht = trees$ht,
                           spcd = trees$spcd,
                           products = cordwood,
                           model = trees$model)
```

```{r, echo = FALSE}
cord_result$logs %>%
  filter(volume_unit == 'cord') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `length (feet)` = length,
            `sed (inches)` = sed,
            `scale (cords)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Prices and validation

`price` is the amount per `price_per` volume units. Omitting `price` leaves the product unpriced and omits log values.
[Prices and bucking](optimal-bucking.html) shows priced specifications.

```{r}
## Check the product table
checked <- specifications

## Count validated products
nrow(checked)
```

## Source rule conversion

`products_from_nvel_rules()` translates explicit source rule limits into products, a stump height, and model auxiliaries.
Continuous length bounds replace source top segment policies. Factor scaling cannot be translated.

```{r}
## Specify the source rule fields to translate
rules <- nvel_rules(even_or_odd = 1,
                    option = 14,
                    maximum_length = 32,
                    minimum_length = 16,
                    primary_top = 6,
                    secondary_top = 3,
                    stump = 1,
                    trim = 0.5,
                    minimum_board_foot_dbh = 6,
                    scribner = 'table')

## Translate the rule record
translated <- products_from_nvel_rules(x = rules)

## Inspect the translated result structure
names(translated)
```
