---
title: 'Defining products'
output:
  rmarkdown::html_vignette:
    md_extensions: -smart
vignette: >
  %\VignetteIndexEntry{Defining products}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      dev.args = list(bg = 'white'), fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

`product()` defines the trees and logs that qualify for a product and its scaling rule. Use explicit specifications when dimensions, eligibility, or measurement rules differ between products. Tree and log diameters enter in inches, and heights, lengths, and trim enter in feet.

## Which trees qualify?

The product name identifies logs in the result and must be unique within a specification. Species, diameter, and age limits keep a product within its accepted tree classes, while pruning requirements limit cuts to the recorded pruned portion. `spcd` limits eligibility to numeric species codes. Leaving it unspecified allows any species with a usable taper equation. Tree diameter limits use outside bark diameter at breast height regardless of the log bark setting.

The minimum tree diameter and age are inclusive. Their maximum limits are exclusive, so adjacent classes can share a boundary without accepting the same tree. Age limits require `age` in `merchandise()`. `requires_pruned = TRUE` requires a supplied `pruned_ht`, and only logs ending within that height qualify.


```{r}
## Attach the calculation and data tools
library(merchandiser)
library(dplyr)
```

```{r}
## Select the tree with a stored taper equation
tree <- example_trees %>%
  filter(tree_id == 5)
```


```{r}
## Define all tree and log eligibility fields explicitly
saw <- product(product = 'saw',  ## product label
               spcd = 202,  ## species code
               min_dbh = 12,  ## inches outside bark
               max_dbh = 30,  ## inches outside bark, excluded
               min_age = 40,  ## years
               max_age = 90,  ## years, excluded
               requires_pruned = TRUE,  ## keep logs within the pruned height
               min_length = 16,  ## feet
               max_length = 32,  ## feet
               length_round = 1,  ## feet
               trim = 0.5,  ## feet
               min_sed = 6,  ## inches
               max_sed = 20,  ## inches
               min_led = 8,  ## inches
               max_led = 30,  ## inches
               inside_bark = TRUE,  ## limits measured inside bark
               max_sweep = 10,  ## percent
               max_logs = 2,  ## logs per available segment
               volume_unit = 'scribner',  ## board feet
               split_scale = FALSE,  ## scale the nominal log without splitting
               round = 'default')  ## use the rule's default diameter rounding
```


```{r}
## Check age, diameter, and pruning eligibility
qualified <- merchandise(tree_id = tree$tree_id,
                         dbh = tree$dbh,
                         ht = tree$ht,
                         spcd = tree$spcd,
                         products = saw,
                         age = 60,
                         pruned_ht = 50)
```


```{r, echo = FALSE}
qualified$logs %>%
  select(product, `length (feet)` = length, `end (feet)` = end_height,
         `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The last selected log ends at `r tail(qualified$logs$end_height, 1)` feet, within the supplied pruning height.


```{r}
## Raise the minimum tree diameter above this tree's diameter
excluded_spec <- saw %>%
  mutate(min_dbh = 21)
```


```{r}
## Recalculate after raising the minimum diameter
excluded <- merchandise(tree_id = tree$tree_id,
                        dbh = tree$dbh,
                        ht = tree$ht,
                        spcd = tree$spcd,
                        products = excluded_spec,
                        age = 60,
                        pruned_ht = 50)
```


```{r}
## Count logs after changing tree eligibility
nrow(excluded$logs)
```

The revised minimum accepts `r nrow(excluded$logs)` logs from this tree because its measured diameter falls below the new threshold.

## Which logs qualify?

`min_length` and `max_length` bound nominal lengths on the half-foot candidate grid. Large-end limits can exclude oversized butt sections, and small-end limits determine whether an upper cut still qualifies. `trim` adds wood above the nominal body, and the physical log must fit within an available section. Both end diameter limits are inclusive and apply at the physical ends, including trim. `min_led` and `max_led` refer to the physical large end, while `min_sed` and `max_sed` refer to the small end.

`max_sweep` sets the largest recorded percentage a product accepts on an overlapping sweep interval. Omitting it imposes no sweep limit. `max_logs` caps selected logs of that product within a segment. This limits repeated assignments to that product before other eligible products are considered. Cull and restriction boundaries start new segments and reset those counts. The [defect guide](recording-defect.html) shows how those intervals enter the calculation.

When rows are combined with `products()`, the first row gets the first opportunity to use each available stem section under the default strategy. Changing that order can change the logs even when every field within each row stays the same.

## How is the accepted log scaled?

`volume_unit` accepts `scribner`, `international`, `doyle`, `cubic`, `cord`, or `green_ton`. The first three return board feet under the named rule. Cubic scale returns cubic feet, and cord scale converts solid volume to stacked volume using `cord_solid_fraction`. Green weight returns short tons including wood and attached bark.

`inside_bark` controls the physical log diameter limits and the cubic or cord volume basis. Board foot rules always use inside bark diameters. Green weight includes attached bark regardless of that setting, so changing bark basis does not turn it into a wood-only mass estimate.

`length_round` rounds the scaling length down. Board foot rules accept whole-foot or even-foot rounding, while other scales accept a positive increment. Cubic, cord, and green weight calculations use nominal length even when the reported scaling length differs. `round` selects `default`, `down`, `nearest`, or `none` for scaling diameter. `split_scale` selects shorter scaling sections for Scribner without adding cuts or affecting other scale units. `none` disables diameter preprocessing, but the source rule can still round internally. These fields affect scale without adding trim to the scaled body.


```{r}
## Hold nominal length and diameter limits fixed for the rounding comparison
rounding_spec <- product(product = 'saw',  ## product label
                         min_length = 16,  ## feet
                         max_length = 16,  ## feet
                         min_sed = 6,  ## inches
                         inside_bark = TRUE,  ## limits measured inside bark
                         trim = 0.5,  ## feet
                         volume_unit = 'scribner',  ## board feet
                         round = 'default')  ## use the rule's diameter rounding
```

```{r}
## Scale the accepted logs with default diameter rounding
rounded <- merchandise(tree_id = tree$tree_id,
                       dbh = tree$dbh,
                       ht = tree$ht,
                       spcd = tree$spcd,
                       products = rounding_spec,
                       model = tree$model)
```

```{r}
## Round scaling diameters down on the same accepted logs
down_spec <- rounding_spec %>%
  mutate(round = 'down')
```

```{r}
## Recalculate with scaling diameters rounded down
rounded_down <- merchandise(tree_id = tree$tree_id,
                            dbh = tree$dbh,
                            ht = tree$ht,
                            spcd = tree$spcd,
                            products = down_spec,
                            model = tree$model)
```

```{r, echo = FALSE}
bind_rows(default = rounded$logs, down = rounded_down$logs, .id = 'rounding') %>%
  select(rounding, log, `scaling diameter (inches)` = scaling_diameter,
         `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
physical_sed <- format(round(rounded$logs$sed[1], 2), trim = TRUE)
```

The first log scales to `r rounded$logs$scale[1]` Scribner board feet under default rounding and `r rounded_down$logs$scale[1]` when rounded down. Its physical small end of `r physical_sed` inches rounds to `r rounded$logs$scaling_diameter[1]` inches under the rule and to `r rounded_down$logs$scaling_diameter[1]` inches when rounded down.

## How are cords specified?

Cords require `cord_solid_fraction`, the solid wood fraction of stacked volume. It must lie strictly between zero and one. The conversion uses the chosen bark basis and divides solid cubic volume by the cord's stacked volume and this fraction.


```{r}
## Convert nominal log bodies to cords with an explicit solid fraction
cordwood <- product(product = 'cordwood',  ## product label
                    min_length = 8,  ## feet
                    max_length = 20,  ## feet
                    min_sed = 3,  ## inches inside bark
                    volume_unit = 'cord',  ## cords
                    cord_solid_fraction = 0.7)  ## solid volume / stacked volume
```


```{r}
## Scale the nominal bodies in cords
cords <- merchandise(tree_id = tree$tree_id,
                     dbh = tree$dbh,
                     ht = tree$ht,
                     spcd = tree$spcd,
                     products = cordwood)
```


```{r, echo = FALSE}
cords$logs %>%
  select(log, `length (feet)` = length, `scale (cords)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```


```{r, echo = FALSE}
cord_text <- format(round(sum(cords$logs$scale), 3), trim = TRUE)
```

The first nominal log body scales to `r format(round(cords$logs$scale[1], 3), trim = TRUE)` cords at the specified solid fraction.

## How are prices applied?

`price` is the amount charged for `price_per` scale units. Currency is supplied by the caller and is not converted. An omitted price leaves the product unpriced. A result includes a value column when any product is priced, with missing values for unpriced products.


```{r}
## Price Scribner scale in dollars per specified board feet
priced <- product(product = 'priced saw',  ## product label
                  min_length = 32,  ## feet
                  max_length = 32,  ## feet
                  min_sed = 12,  ## inches inside bark
                  trim = 0.5,  ## feet
                  volume_unit = 'scribner',  ## board feet
                  price = 900,  ## dollars
                  price_per = 1000)  ## board feet
```


```{r}
## Apply the stated price to accepted logs
valued <- merchandise(tree_id = tree$tree_id,
                      dbh = tree$dbh,
                      ht = tree$ht,
                      spcd = tree$spcd,
                      products = priced)
```


```{r, echo = FALSE}
valued$logs %>%
  select(log, `scale (Scribner board feet)` = scale, `value (dollars)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The first log contributes `r valued$logs$value[1]` dollars from its scale multiplied by `price / price_per`. Prices affect log selection only when `strategy = 'optimize'` is requested.
