---
title: 'Fitting and completing heights'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting and completing heights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`fit_height()` fits height from diameter by species. A `group` vector adds supported group effects.
`complete_heights()` retains finite heights and fills missing or nonfinite values.

## Fit measured heights

```{r}
## Select retained height measurements
measured <- example_trees_pnw %>%
  filter(!is.na(ht_observed))
```

```{r}
## Fit height by species with plot effects
fit <- fit_height(dbh = measured$dbh,
                  ht = measured$ht_observed,
                  spcd = measured$spcd,
                  group = measured$plot)
```

```{r, echo = FALSE}
fit$fixed_effects %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`pooled` identifies species using the pooled relationship. Species with too few measurements or unsuccessful separate fits use that relationship.
Available forms are `chapman_richards`, `curtis`, `wykoff`, `naslund`, and `schumacher`.

```{r}
## Draw measured heights and fitted curves
plot(fit)
```

## Complete inventory heights

```{r}
## Retain observed heights and fill the missing values
trees <- example_trees_pnw %>%
  mutate(ht = complete_heights(dbh = dbh,
                               ht = ht_observed,
                               spcd = spcd,
                               group = plot,
                               fit = fit))
```

```{r, echo = FALSE}
trees %>%
  filter(is.na(ht_observed)) %>%
  select(tree, species, dbh, ht_observed, ht) %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Pass a fitted object to reuse the same model across data groups.
With `fit = NULL`, completion fits from the supplied observations.

## Predict heights

`predict_height()` predicts every supplied tree. `re_form = 'population'` omits fitted group effects.
Groups absent from the fit use its population relationship. Species absent from the fit receive missing predictions.

```{r}
## Predict heights for the complete shipped tree list
predictions <- predict_height(fit = fit,
                              dbh = example_trees$dbh,
                              spcd = example_trees$spcd,
                              re_form = 'population',
                              interval = TRUE,
                              seed = 7)
```

```{r, echo = FALSE}
predictions %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

With `interval = TRUE`, the result contains `fit`, `lower`, and `upper`.
Simulated limits include group variation and residual error with coefficients held fixed.
They exclude coefficient uncertainty and inventory sampling uncertainty.
