---
title: 'Fitting and completing heights'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting and completing heights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

`fit_height()` fits height by species, with optional plot adjustments.
`complete_heights()` fills missing heights for subsequent stem calculations.

```{r}
## Blank heights whose shipped origin is predicted
trees <- example_trees_pnw %>%
  mutate(ht = if_else(condition = ht_status == 'predicted',
                      true = NA_real_,
                      false = ht))

## Retain measured heights for fitting
measured <- trees %>%
  filter(ht_status == 'measured')

## Fit height with plot adjustments
fit <- fit_height(dbh = measured$dbh,
                  ht = measured$ht,
                  spcd = measured$spcd,
                  group = measured$plot)
```

<details>
<summary>Measured heights and fitted curves</summary>

```{r, echo = FALSE}
fit$data %>%
  rename(`diameter (inches)` = dbh, `measured height (feet)` = ht) %>%
  knitr::kable(digits = 3, row.names = FALSE)
curves <- fit$data %>%
  group_by(spcd) %>%
  reframe(dbh = seq(min(dbh), max(dbh), length.out = 100)) %>%
  mutate(height = predict_height(fit = fit, dbh = dbh, spcd = spcd,
                                 re_form = 'population'))
curves %>%
  rename(`diameter (inches)` = dbh, `predicted height (feet)` = height) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

</details>

The curves show population predictions against measured heights, separately by species.

```{r, echo = FALSE, fig.width = 9}
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
plot(fit)
```

Available forms are `chapman_richards`, `curtis`, `wykoff`, `naslund`, and `schumacher`.
Species below `min_n`, or with unsuccessful separate fits, use the pooled relationship.

```{r}
## Refill the blanked heights
completed <- trees %>%
  mutate(ht = complete_heights(dbh = dbh,
                               ht = ht,
                               spcd = spcd,
                               group = plot,
                               fit = fit))

## Compare refilled heights with the shipped values
comparison <- completed %>%
  mutate(shipped_ht = example_trees_pnw$ht,
         difference = ht - shipped_ht) %>%
  filter(ht_status == 'predicted')
```

```{r, echo = FALSE}
comparison %>%
  slice_head(n = 6) %>%
  transmute(tree_id = tree_id, `refilled height (feet)` = ht,
            `shipped height (feet)` = shipped_ht, `difference (feet)` = difference) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Check agreement across every refilled height
comparison %>%
  summarize(maximum_difference_feet = max(abs(difference)))
```

## Predict heights

`predict_height()` predicts all supplied rows. Population predictions omit plot adjustments.
Unseen groups use population predictions. Unseen species return missing heights.

```{r}
## Predict heights and intervals for the small tree list
predictions <- predict_height(fit = fit,
                              dbh = example_trees$dbh,
                              spcd = example_trees$spcd,
                              re_form = 'population',
                              interval = TRUE,
                              seed = 7)
```

```{r, echo = FALSE}
predictions %>%
  rename(`height (feet)` = fit, `lower (feet)` = lower, `upper (feet)` = upper) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Intervals include group variation and residual error with fixed coefficients. They exclude coefficient and inventory sampling uncertainty.
