---
title: 'Get started'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`merchandise()` selects logs from tree measurements and an ordered product table.
It returns log dimensions, scaled quantities, tree totals, residual intervals, and recorded settings.
The shipped tree inventories and product specifications are synthetic examples.

## Tree and products

Default inputs use inches for diameter and feet for height.

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load the example specifications
specifications <- example_products(name = 'douglas_fir')
```

```{r, echo = FALSE}
specifications %>%
  mutate(lengths = vapply(lengths, paste, character(1), collapse = ', ')) %>%
  select(product, lengths, min_length, max_length, min_sed, trim, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`volume_unit` specifies each product's measurement. The example prices use United States dollars (`USD`).

## Logs and result tables

```{r}
## Select logs from the tree
result <- merchandise(dbh = tree$dbh,
                      ht = tree$ht,
                      model = tree$model,
                      species = tree$species,
                      id = tree$tree,
                      products = specifications,
                      currency = 'USD',
                      status = TRUE)
```

```{r, echo = FALSE}
result$logs %>%
  select(log, product, nominal_length, sed_ib, net_scale, scale_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`logs` holds the selected cuts. `trees` retains every input tree, including trees without logs.
`net_scale` uses the accompanying `scale_unit`.

```{r}
## Select tree totals
totals <- result$trees %>%
  select(id, log_count, log_net_cubic_ib, net_value, status)
```

```{r, echo = FALSE}
totals %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Draw the selected logs
plot(result)
```

`residuals` records stem intervals outside the logs. `scales` contains product measurements and any additional reports.
`defect_accounting` records deductions, and `values` records priced quantities.


## Inventory

`ht_simulated` supplies complete example heights. [complete_heights()](fill-missing-heights.html) fills missing `ht_observed` values from retained measurements.

```{r}
## Use the complete simulated heights and retain tree identifiers
trees <- example_trees_pnw %>%
  mutate(id = tree, ht = ht_simulated)
```

```{r}
## Load the inventory product specifications
specifications <- example_products(name = 'pnw')
```

```{r}
## Apply the product table to the inventory
result <- merchandise(dbh = trees$dbh,
                      ht = trees$ht,
                      species = trees$species,
                      id = trees$id,
                      products = specifications,
                      status = TRUE)
```

```{r}
## Expand the inventory by species
stand <- stand_table(x = result,
                     trees = trees,
                     expansion = trees$expansion_factor,
                     by = 'species')
```

```{r, echo = FALSE}
stand %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Expansion factors are used as supplied. These example weights do not represent a sampling design.
`merchandising_status` separates valid quantities, failed calculations, and inventory trees absent from the result.
