---
title: 'Get started'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product order in `products()` sets cutting priority. `merchandise()` returns logs, residuals, status, and assumptions for checking product yield.

```{r}
## Load the packages
library(merchandiser)

## Load data verbs
library(dplyr)

## Select example trees
trees <- example_trees %>%
  slice_tail(n = 3)
```

## Products

The products match `data-raw/example-products.R`.

```{r}
## Define export specifications
export <- product(product = 'export',  ## product name, unitless
                  spcd = c(202, 263),  ## species codes, unitless
                  min_dbh = 12,  ## inches
                  min_length = 32,  ## feet
                  max_length = 40,  ## feet
                  trim = 1,  ## feet
                  min_sed = 12,  ## inches
                  inside_bark = TRUE,  ## logical, unitless
                  volume_unit = 'scribner',  ## board feet
                  split_scale = FALSE)  ## logical, unitless

## Define domestic saw specifications
domestic_saw <- product(product = 'domestic_saw',  ## product name, unitless
                        spcd = c(202, 263),  ## species codes, unitless
                        min_dbh = 10,  ## inches
                        min_length = 16,  ## feet
                        max_length = 40,  ## feet
                        trim = 1,  ## feet
                        min_sed = 6,  ## inches
                        inside_bark = TRUE,  ## logical, unitless
                        volume_unit = 'scribner',  ## board feet
                        split_scale = FALSE)  ## logical, unitless

## Define pulp specifications
pulp <- product(product = 'pulp',  ## product name, unitless
                spcd = c(202, 263),  ## species codes, unitless
                min_dbh = 5,  ## inches
                min_length = 8,  ## feet
                max_length = 40,  ## feet
                trim = 0.5,  ## feet
                min_sed = 3,  ## inches
                inside_bark = TRUE,  ## logical, unitless
                volume_unit = 'green_ton')  ## green short tons

## Combine products in cutting priority order
specifications <- products(export = export,
                           domestic_saw = domestic_saw,
                           pulp = pulp)

## Cut and scale the trees
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = specifications,
                      model = trees$model)
```

## Logs

Log numbers restart within each tree. Diameters describe physical log ends on the recorded bark basis.

```{r, echo = FALSE}
result$logs %>%
  filter(volume_unit == 'scribner') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `start_height (feet)` = start_height,
            `end_height (feet)` = end_height,
            `length (feet)` = length,
            `scaling_length (feet)` = scaling_length,
            `sed (inches)` = sed,
            `led (inches)` = led,
            `scaling_diameter (inches)` = scaling_diameter,
            `inside_bark` = inside_bark,
            `scale (Scribner board feet)` = scale,
            `volume_unit` = volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
result$logs %>%
  filter(volume_unit == 'green_ton') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `start_height (feet)` = start_height,
            `end_height (feet)` = end_height,
            `length (feet)` = length,
            `scaling_length (feet)` = scaling_length,
            `sed (inches)` = sed,
            `led (inches)` = led,
            `scaling_diameter (inches)` = scaling_diameter,
            `inside_bark` = inside_bark,
            `scale (green short tons)` = scale,
            `volume_unit` = volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Residuals

Residual intervals identify stem sections outside the scaled log bodies.

```{r, echo = FALSE}
result$residuals %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Status

A clean run has an empty status table.

```{r}
## Show the clean status table
result$status
```

## Assumptions

`value` uses the accompanying `unit`.

```{r, echo = FALSE}
result$assumptions %>%
  mutate(source = if_else(assumption == 'species_model',
                          'pinned source library', source)) %>%
  rename(`value (unit column)` = value, `source (abbreviated)` = source) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Drawing

<details>
<summary>Drawing measurements</summary>

```{r, echo = FALSE}
drawing_trees <- trees %>%
  slice_head(n = 1)
drawing_rows <- match(drawing_trees$tree_id, result$call$tree_id)
drawing_models <- result$call$model[drawing_rows]
drawing_profiles <- stem_profile(
  tree_id = drawing_trees$tree_id,
  dbh = drawing_trees$dbh,
  ht = drawing_trees$ht,
  spcd = drawing_trees$spcd,
  model = drawing_models,
  step = 0.5,
  from = 0
)
drawing_profiles %>%
  transmute(tree_id, `height (feet)` = h, `radius inside bark (inches)` = dib / 2) %>%
  knitr::kable(digits = 4, row.names = FALSE)
```

</details>

The drawing locates the selected products and residual sections along the first tree.

```{r, echo = FALSE}
plot(result, tree_id = trees$tree_id[1], ylab = 'Stem radius (inches)')
```

`call` retains inputs for replaying the calculation.
