---
title: 'Identifiers and defaults'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identifiers and defaults}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Equation identifiers select taper implementations, and species codes select equation scope and properties.
Stored assignments and registry records support repeating a calculation.

## Stored equations

```{r}
## Inspect stored species and model assignments
assignments <- example_trees %>%
  distinct(spcd, model)
```

```{r, echo = FALSE}
assignments %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Check the stored assignments
check_taper_models(model = assignments$model, spcd = assignments$spcd)
```

An empty diagnosis table means the assignments passed. Explicit `model` takes precedence over `taper_map` and species defaults.

## Species defaults

Automatic selections are recorded by `merchandise()`. The result retains the selected equation for every tree.

```{r}
## Define saw specifications
saw <- product(product = 'domestic_saw',  ## product name, unitless
               min_length = 16,  ## feet
               max_length = 40,  ## feet
               trim = 1,  ## feet
               min_sed = 6,  ## inches
               volume_unit = 'scribner')  ## Scribner board feet

## Cut and scale the trees
result <- merchandise(tree_id = example_trees$tree_id,
                      dbh = example_trees$dbh,
                      ht = example_trees$ht,
                      spcd = example_trees$spcd,
                      products = saw)
```

```{r, echo = FALSE}
result$assumptions %>%
  filter(assumption == 'species_model') %>%
  select(tree_id, spcd, model) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Species properties

```{r}
## Select the example species records
species <- species_reference %>%
  filter(spcd %in% assignments$spcd)
```

```{r, echo = FALSE}
species %>%
  select(spcd, common, scientific, `bark ratio (unitless)` = bark_ratio) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`species_reference` stores names, wood density, and bark ratios. Calculation calls accept numeric species codes.
County and coordinate lookups select optional [biomass coefficients](biomass-carbon.html).

## Registry record

```{r}
## Record the current registry
manifest <- taper_manifest()

```

```{r, echo = FALSE}
manifest %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The manifest records registered identifiers, owning packages, versions, and registration order.
It excludes coefficients and source identifiers resolved without explicit registration. Save fitted objects or coefficient vectors separately.
