---
title: 'Identifiers and defaults'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identifiers and defaults}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

Equation identifiers select a taper implementation. Species identifiers select species scope and property records.
Forest Service region, forest, and district fields are agency codes used by the source default lookup.

## Stored equations

```{r}
## Inspect equations stored with the shipped tree list
assignments <- example_trees %>%
  distinct(spcd, species, model)
```

```{r, echo = FALSE}
assignments %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Check the stored species and equation assignments
check_taper_models(id = assignments$model, spcd = assignments$spcd)
```

An empty diagnosis table means the assignment checks passed.
An explicit `model` takes precedence over `taper_map` and automatic defaults.

## Source defaults

`nvel_default_equation()` applies the pinned lookup to agency codes already present in the example inventory.

```{r}
## Resolve equations for the shipped agency-code combinations
defaults <- example_trees_pnw %>%
  distinct(region, forest, district, spcd) %>%
  mutate(model = nvel_default_equation(region = region,
                                       forest = forest,
                                       district = district,
                                       spcd = spcd))
```

```{r, echo = FALSE}
defaults %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Attach the resolved equations to inventory rows
trees <- example_trees_pnw %>%
  left_join(defaults, by = c('region', 'forest', 'district', 'spcd'))
```

```{r, echo = FALSE}
trees %>%
  select(tree, species, model) %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`nvel_from_fia_code()` translates Forest Inventory and Analysis inventory equation codes into source identifiers.
Its `geosub` and `primary_top` arguments participate in source substitutions.

## Species and ecological codes

```{r}
## Resolve species records for the example tree list
species <- species_lookup(x = assignments$spcd, from = 'spcd')
```

```{r, echo = FALSE}
species %>%
  select(spcd, common, scientific) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`species_reference` stores names, wood density, and bark ratios.
`nsvb_division()` returns county-predominant ecological codes from state and county codes.
`nsvb_division_xy()` uses coordinates and packaged ecological boundaries for location-based assignment.

## Registry record

```{r}
## Save the current equation registrations
manifest <- taper_manifest()

## Compare the saved record with the current registry
check_taper_manifest(manifest = manifest)
```

```{r, echo = FALSE}
manifest %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The manifest records registered identifiers, owning packages, versions, and registration order.
It excludes local coefficients and source identifiers resolved without explicit registration.
Save fitted objects or coefficient vectors separately.
