---
title: 'Taper equation methods'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taper equation methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Shipped taper equations follow the pinned National Volume Estimator Library source revision.
Source and agreement tables identify implementations available for stem calculations.

```{r}
## Inspect the pinned source revision
nvel_source_revision()
```

## Source routines

Coefficient inventories retain source filenames, line locations, and revisions.

```{r, echo = FALSE}
families <- c(
  'Flewelling variable-shape profiles',
  'Clark source profiles',
  'Alaska and Mathis source profiles',
  'Smaller source profile families',
  'National scale volume and biomass profiles',
  'Identifier and merchandising support'
)

flewelling_files <- paste(
  c(
    'f_west.f',
    'f_ingy.f',
    'f_alaska.f',
    'f_other.f',
    'sf_2pt.f',
    'sf_2pth.f',
    'sf_3pt.f',
    'sf_3z.f',
    'sf_corr.f',
    'sf_dfz.f',
    'sf_ds.f',
    'sf_hs.f',
    'sf_shp.f',
    'sf_taper.f',
    'sf_yhat.f',
    'sf_yhat3.f',
    'sf_zero.f'
  ),
  collapse = ', '
)

clark_files <- paste(
  c(
    'r8clkdib.f',
    'r8dib.f',
    'r8dib.inc',
    'r8clkcoef.inc',
    'r8cfo.inc',
    'r8clist.inc',
    'r8vlist.f',
    'r8vlist.inc',
    'r8init.f',
    'r8prep.f',
    'r8vol.f',
    'r8vol1.f',
    'r8vol2.f',
    'r9clarkdib.f',
    'r9clark.f',
    'r9coeff.inc',
    'r9init.f',
    'r9vol.f',
    'r9logs.f',
    'clkcoef_mod.f',
    'calcdia.f',
    'ht2topd.f',
    'profile.f',
    'volinit.f',
    'voleqdef.f'
  ),
  collapse = ', '
)

alaska_files <- paste(
  c(
    'r10tap.f',
    'r10tapo.f',
    'r10d2h.f',
    'r10vol.f',
    'r10vol1.f',
    'r10volo.f',
    'r4d2h.f',
    'r4vol.f',
    'profile.f',
    'calcdia.f',
    'ht2topd.f'
  ),
  collapse = ', '
)

source_files <- paste(
  c(
    'r1tap.f',
    'r2tap.f',
    'r5tap.f',
    'r12tap.f',
    'stump.f',
    'blmtap.f',
    'blmvol.f',
    'formclas.f',
    'r6vol3.f'
  ),
  collapse = ', '
)

national_files <- paste(
  c(
    'nsvb.f',
    'tables1.inc through tables11.inc',
    'dist_ecoprov.inc',
    'regndftdata.inc',
    'scrib.f'
  ),
  collapse = ', '
)

identifier_files <- paste(c('voleqdef.f', 'fiaeq2nveleq.for', 'mrules.f'),
  collapse = ', '
)

family_sources <- data.frame(
  family = families,
  source_files = c(
    flewelling_files,
    clark_files,
    alaska_files,
    source_files,
    national_files,
    identifier_files
  )
)

knitr::kable(family_sources, row.names = FALSE)
```

National biomass diameter and cumulative volume calculations use distinct fitted relationships.
[Stem calculations](stem-model-overview.html) use the capabilities of the selected equation.

## Identifier fields

Field positions depend on the equation family.

```{r, echo = FALSE}
profile_id <- example_trees$model[1]
national_id <- 'NVB0210202'
```

```{r, echo = FALSE}
profile_parts <- data.frame(
  layout = 'source profile',
  field = c(
    'geographic equation selection',
    'equation mnemonic',
    'family-specific variant',
    'species suffix'
  ),
  positions = c('1:3', '4:6', '7', '8:10'),
  value = c(
    substr(profile_id, 1, 3),
    substr(profile_id, 4, 6),
    substr(profile_id, 7, 7),
    substr(profile_id, 8, 10)
  )
)

national_parts <- data.frame(
  layout = 'national scale',
  field = c(
    'system marker',
    'mountain indicator or zero',
    'ecological division',
    'species suffix'
  ),
  positions = c('1:3', '4', '5:7', '8:10'),
  value = c(
    substr(national_id, 1, 3),
    substr(national_id, 4, 4),
    substr(national_id, 5, 7),
    substr(national_id, 8, 10)
  )
)

identifier_parts <- bind_rows(profile_parts, national_parts)

knitr::kable(identifier_parts, row.names = FALSE)
```

`check_taper_models()` checks equation availability and species scope. [Identifiers and defaults](identifiers-defaults.html) shows stored assignments.

## Compatibility behavior

The default compatibility option is `port`. The `nvel` option reproduces source behavior for the listed differences.

```{r}
## Inspect the compatibility option
getOption(x = 'merchandiser.compat', default = 'port')
```

```{r, echo = FALSE}
area <- c(
  'FIAEQ2NVELEQ geographic substitution',
  'Flewelling SF_HS inverse',
  'Clark 834CLKE110 inverse',
  'Clark R9LOGS error 12',
  'R10 height selection and cedar equality',
  'Smaller-family inverse search',
  'R12 height selection',
  'National scale CALCDIA outside-bark diameter',
  'National scale Table S10 carbon fraction'
)

port <- c(
  'applies the intended trimmed geographic substitutions',
  'inverts the displayed profile and selects its highest crossing',
  'returns the highest actual crossing',
  'retains a computable total in the internal volume record',
  'inverts the registered profile consistently',
  'returns the precise numeric crossing',
  'inverts the public profile',
  'returns the outside-bark diameter computed internally',
  'returns the fraction used by the biomass calculation'
)

nvel <- c(
  'reproduces fixed-width source string semantics',
  'reproduces the source bark-height argument and crossing selection',
  'reproduces the source two-root addition',
  'reproduces the source error and cleared record',
  'retains source height results and the cedar equality rule',
  'reproduces the source rounded search',
  'retains the source result for an unselected calculation',
  'reproduces the unassigned source output slot',
  'returns the raw source-table fraction'
)

compatibility_differences <- data.frame(area = area, port = port, nvel = nvel)

knitr::kable(compatibility_differences, row.names = FALSE)
```

## Fixture agreement

The saved table compares double and single precision source builds with package calculations. Tests enforce tolerances and named exclusions.

```{r, echo = FALSE}
agreement <- read.csv(system.file('extdata',
  'oracle-agreement.csv',
  package = 'merchandiser'
))
```

```{r, echo = FALSE}
agreement %>%
  distinct(precision, compat) %>%
  arrange(precision, compat) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
agreement %>%
  mutate(output = recode(output,
                         DIB = 'diameter inside bark',
                         DOB = 'diameter outside bark',
                         `height at DIB` = 'height at inside bark diameter',
                         `cedar equality height at DIB` =
                           'cedar equality height at inside bark diameter')) %>%
  mutate(
    max_relative_difference = format(max_relative_difference, scientific = TRUE, digits = 12),
    tolerance = format(tolerance, scientific = TRUE, digits = 12)
  ) %>%
  rename(`rows compared (count)` = rows_compared,
         `rows within tolerance (count)` = rows_within_tolerance,
         `maximum relative difference (unitless)` = max_relative_difference,
         `tolerance (unitless)` = tolerance, `exclusions (names and counts)` = exclusions) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Full source fixtures are supplied separately to the test suite.
