---
title: 'Taper equation methods'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taper equation methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

The shipped equations follow a pinned National Volume Estimator Library source revision.
`nvel_source_revision()` returns that revision and attributes identifying the repository and fixture release.

`DIB` denotes diameter inside bark, and `DOB` denotes diameter outside bark.

```{r}
## Report the equation source revision
nvel_source_revision()
```

## Source routines

Coefficient inventories retain source files, line locations, and revision identifiers.


```{r, echo = FALSE}
families <- c('Flewelling variable-shape profiles',
              'Clark regional profiles',
              'Alaska and Mathis regional profiles',
              'Smaller regional profile families',
              'National scale volume and biomass profiles',
              'Identifier and merchandising support')

flewelling_files <- paste(c('f_west.f',
                            'f_ingy.f',
                            'f_alaska.f',
                            'f_other.f',
                            'sf_2pt.f',
                            'sf_2pth.f',
                            'sf_3pt.f',
                            'sf_3z.f',
                            'sf_corr.f',
                            'sf_dfz.f',
                            'sf_ds.f',
                            'sf_hs.f',
                            'sf_shp.f',
                            'sf_taper.f',
                            'sf_yhat.f',
                            'sf_yhat3.f',
                            'sf_zero.f'),
                          collapse = ', ')

clark_files <- paste(c('r8clkdib.f',
                       'r8dib.f',
                       'r8dib.inc',
                       'r8clkcoef.inc',
                       'r8cfo.inc',
                       'r8clist.inc',
                       'r8vlist.f',
                       'r8vlist.inc',
                       'r8init.f',
                       'r8prep.f',
                       'r8vol.f',
                       'r8vol1.f',
                       'r8vol2.f',
                       'r9clarkdib.f',
                       'r9clark.f',
                       'r9coeff.inc',
                       'r9init.f',
                       'r9vol.f',
                       'r9logs.f',
                       'clkcoef_mod.f',
                       'calcdia.f',
                       'ht2topd.f',
                       'profile.f',
                       'volinit.f',
                       'voleqdef.f'),
                     collapse = ', ')

alaska_files <- paste(c('r10tap.f',
                        'r10tapo.f',
                        'r10d2h.f',
                        'r10vol.f',
                        'r10vol1.f',
                        'r10volo.f',
                        'r4d2h.f',
                        'r4vol.f',
                        'profile.f',
                        'calcdia.f',
                        'ht2topd.f'),
                      collapse = ', ')

regional_files <- paste(c('r1tap.f',
                          'r2tap.f',
                          'r5tap.f',
                          'r12tap.f',
                          'stump.f',
                          'blmtap.f',
                          'blmvol.f',
                          'formclas.f',
                          'r6vol3.f'),
                        collapse = ', ')

national_files <- paste(c('nsvb.f',
                          'tables1.inc through tables11.inc',
                          'dist_ecoprov.inc',
                          'regndftdata.inc',
                          'scrib.f'),
                        collapse = ', ')

identifier_files <- paste(c('voleqdef.f', 'fiaeq2nveleq.for', 'mrules.f'),
                          collapse = ', ')

family_sources <- data.frame(family = families,
                             source_files = c(flewelling_files,
                                              clark_files,
                                              alaska_files,
                                              regional_files,
                                              national_files,
                                              identifier_files))

knitr::kable(family_sources, row.names = FALSE)
```

National biomass diameter and cumulative-volume calculations use distinct fitted relationships.
[Stem functions](stem-model-overview.html) accept equation identifiers and measurement vectors.
[fit_taper()](methods-fitting-taper.html) estimates local coefficients.

## Identifier fields

The source identifiers use fixed character positions. Field meanings depend on the family.

```{r, echo = FALSE}
regional_id <- nvel_default_equation(region = 1, forest = 3, district = 0, spcd = 202)
national_id <- 'NVB0210202'
```

```{r, echo = FALSE}
regional_parts <- data.frame(layout = 'regional profile',
                             field = c('geographic equation selection',
                                       'equation mnemonic',
                                       'family-specific variant',
                                       'species suffix'),
                             positions = c('1:3', '4:6', '7', '8:10'),
                             value = c(substr(regional_id, 1, 3),
                                       substr(regional_id, 4, 6),
                                       substr(regional_id, 7, 7),
                                       substr(regional_id, 8, 10)))

national_parts <- data.frame(layout = 'national scale',
                             field = c('system marker',
                                       'mountain indicator or zero',
                                       'ecological division',
                                       'species suffix'),
                             positions = c('1:3', '4', '5:7', '8:10'),
                             value = c(substr(national_id, 1, 3),
                                       substr(national_id, 4, 4),
                                       substr(national_id, 5, 7),
                                       substr(national_id, 8, 10)))

identifier_parts <- bind_rows(regional_parts, national_parts)

knitr::kable(identifier_parts, row.names = FALSE)
```

`check_taper_models()` checks resolved identifiers and species scope.
[nvel_default_equation()](identifiers-defaults.html) resolves identifiers from agency and species codes.

## Compatibility behavior

`options(merchandiser.compat = 'port')` is the default.
`options(merchandiser.compat = 'nvel')` selects source-compatible calculations.
Restore the previous process option after a source comparison.

```{r, echo = FALSE}
area <- c('FIAEQ2NVELEQ geographic substitution',
          'Flewelling SF_HS inverse',
          'Clark 834CLKE110 inverse',
          'Regional Clark R9LOGS error 12',
          'Regional R10 height selection and cedar equality',
          'Smaller-family inverse search',
          'Regional R12 height selection',
          'National scale CALCDIA outside-bark diameter',
          'National scale Table S10 carbon fraction')

port <- c('applies the intended trimmed geographic substitutions',
          'inverts the displayed profile and selects its highest crossing',
          'returns the highest actual crossing',
          'retains a computable total in the internal volume record',
          'inverts the registered profile consistently',
          'returns the precise numeric crossing',
          'inverts the public profile',
          'returns the outside-bark diameter computed internally',
          'returns the fraction used by the biomass calculation')

nvel <- c('reproduces fixed-width source string semantics',
          'reproduces the source bark-height argument and crossing selection',
          'reproduces the source two-root addition',
          'reproduces the source error and cleared record',
          'retains source height results and the cedar equality rule',
          'reproduces the source rounded search',
          'retains the source result for an unselected calculation',
          'reproduces the unassigned source output slot',
          'returns the raw source-table fraction')

compatibility_differences <- data.frame(area = area, port = port, nvel = nvel)

knitr::kable(compatibility_differences, row.names = FALSE)
```

## Fixture agreement

Differential tests compare package results with double- and single-precision source builds.
They enforce tolerances and named exclusions for each family and operation.

```{r, echo = FALSE}
agreement <- read.csv(system.file('extdata',
                                  'oracle-agreement.csv',
                                  package = 'merchandiser'))
```

```{r, echo = FALSE}
agreement %>%
  distinct(precision, compat) %>%
  arrange(precision, compat) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
agreement %>%
  mutate(max_relative_difference = format(max_relative_difference, scientific = TRUE, digits = 12),
         tolerance = format(tolerance, scientific = TRUE, digits = 12)) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The saved table ships with the package. Full source fixtures are supplied separately to the test suite.
