---
title: 'Fitting taper equations'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting taper equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`fit_taper()` estimates coefficients for compiled taper forms from repeated stem measurements.
`as_taper_model()` converts a fit into a model accepted by the stem registry.

## Fitting data and forms

Required columns are `tree`, `dbh`, `ht`, `h`, and `dib`.
Tree diameter and total height must remain constant within each tree.
The shipped `example_stem_measurements` contains modeled diameters from `example_trees`, not independent field measurements.

```{r}
## Select the shipped fitting measurements
measurements <- example_stem_measurements %>%
  select(tree, dbh, ht, h, dib)
```

```{r, echo = FALSE}
measurements %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
forms <- dplyr::tribble(~form,
                        ~source,
                        ~structure,
                        'kozak_1988',
                        'Kozak (1988), doi:10.1139/x88-213',
                        'Variable exponent with reference point',
                        'kozak_2002',
                        'Kozak (2004), doi:10.5558/tfc80507-4',
                        'Variable exponent with tree and height terms',
                        'max_burkhart',
                        'Max and Burkhart (1976), doi:10.1093/forestscience/22.3.283',
                        'Segmented polynomial with fitted join points')
knitr::kable(forms)
```

The package supplies equation forms without published coefficient vectors.
`units` describes the input measurements. Fitted coefficients use centimeters and meters.

## Fit and inspect

```{r}
## Fit the segmented taper form
fit <- fit_taper(data = measurements,
                 form = 'max_burkhart',
                 units = 'imperial')
```

```{r, echo = FALSE}
fit$fit_statistics$overall %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Draw fitted diameters and residuals
plot(fit)
```

The fit stores coefficients, fitted values, residuals, and fit statistics.
`rmse` is root mean squared error, and `bias` is mean observed minus fitted diameter. Both use centimeters.

`summary()` reports fit statistics and coefficients. `predict()` evaluates new measurement rows.

Supplying `group` requests a mixed-effects fit.
The random effect acts on the Kozak scale parameter or the segmented form's `b1` parameter.
Predictions use population coefficients when a group was absent from the fitting data.

```{r}
## Predict diameters at the supplied measurement heights
predicted <- measurements %>%
  mutate(fitted_dib = predict(object = fit,
                              newdata = measurements,
                              units = 'imperial'))
```

```{r, echo = FALSE}
predicted %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Register fitted coefficients

`as_taper_model()` creates a model from population coefficients. `register_taper_model()` makes it available for the current session.
An explicit `bark_ratio` enables outside-bark calculations for the fitted model.

```{r}
## Create and register a model from the fitted coefficients
local_model <- as_taper_model(x = fit,
                              id = 'example.segmented',
                              bark_ratio = 0.9)

## Make the model available for stem calculations
register_taper_model(model = local_model)
```

```{r}
## Calculate volume with the fitted model
volumes <- example_trees %>%
  transmute(tree,
            cubic = stem_volume(dbh = dbh,
                                ht = ht,
                                model = local_model$id))
```

```{r, echo = FALSE}
volumes %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Validate and remove the session model
validate_taper_model(x = local_model)

## Remove the example registration
unregister_taper_model(id = local_model$id)
```

`taper_model_from_coefficients()` constructs the same compiled model type from a named coefficient vector.
`new_taper_model()` constructs a model from R callbacks.
