---
title: 'Optimal bucking'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimal bucking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

`merchandise()` can follow product order or maximize log value. These illustrative prices compare both strategies on the Pacific Northwest trees.

```{r}
## Retain the complete example list
trees <- example_trees_pnw
```

```{r}
## Define domestic saw specifications
domestic_saw <- product(product = 'domestic_saw',  ## product name, unitless
                        min_length = 16,  ## feet
                        max_length = 40,  ## feet
                        trim = 1,  ## feet
                        min_sed = 6,  ## inches
                        volume_unit = 'scribner',  ## Scribner board feet
                        price = 700,  ## dollars
                        price_per = 1000)  ## board feet
```

```{r}
## Define pulp specifications
pulp <- product(product = 'pulp',  ## product name, unitless
                min_length = 8,  ## feet
                max_length = 40,  ## feet
                trim = 0.5,  ## feet
                min_sed = 3,  ## inches
                volume_unit = 'green_ton',  ## green short tons
                price = 30,  ## dollars
                price_per = 1)  ## green short tons
```

```{r}
## Combine products in cutting priority order
specifications <- products(domestic_saw = domestic_saw,
                           pulp = pulp)
```

```{r}
## Cut and scale the trees
cascade <- merchandise(tree_id = trees$tree_id,
                       dbh = trees$dbh,
                       ht = trees$ht,
                       spcd = trees$spcd,
                       products = specifications,
                       age = trees$age,
                       strategy = 'cascade')
```

```{r}
## Cut and scale the trees
optimized <- merchandise(tree_id = trees$tree_id,
                         dbh = trees$dbh,
                         ht = trees$ht,
                         spcd = trees$spcd,
                         products = specifications,
                         age = trees$age,
                         strategy = 'optimize')
```

```{r, echo = FALSE}
bind_rows(cascade = cascade$logs, optimized = optimized$logs, .id = 'strategy') %>%
  filter(tree_id %in% c(7, 14, 28)) %>%
  select(strategy, tree_id, log, product, start_height, length, value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The left drawings show cascade cuts. The right drawings show optimized cuts on the same stems.

```{r, echo = FALSE, fig.width = 16, fig.height = 9}
par(mfrow = c(3, 2), mar = c(4, 4, 1, 1))
for (id in c(7, 14, 28)) {
  plot(cascade, tree_id = id, main = '', ylab = 'Stem radius (inches)')
  plot(optimized, tree_id = id, main = '', ylab = 'Stem radius (inches)')
}
```

```{r}
## Sum cascade log values by tree
cascade_values <- cascade$logs %>%
  group_by(tree_id) %>%
  summarize(cascade_dollars = sum(value))

## Sum optimized log values by tree
optimized_values <- optimized$logs %>%
  group_by(tree_id) %>%
  summarize(optimized_dollars = sum(value))

## Calculate each tree's value gain
gain <- trees %>%
  select(tree_id = tree_id) %>%
  left_join(y = cascade_values,
            by = 'tree_id') %>%
  left_join(y = optimized_values,
            by = 'tree_id') %>%
  mutate(cascade_dollars = coalesce(cascade_dollars, 0),
         optimized_dollars = coalesce(optimized_dollars, 0),
         gain_dollars = optimized_dollars - cascade_dollars)
```

```{r, echo = FALSE}
gain %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Sum values and gains across the list
total <- gain %>%
  summarize(cascade_dollars = sum(cascade_dollars),
            optimized_dollars = sum(optimized_dollars),
            gain_dollars = sum(gain_dollars))
```

```{r, echo = FALSE}
total %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Mapped defects

The same trees now carry `example_defects_pnw`. Restrictions, culls, sweep, and broken tops change the available cuts.

```{r}
## Cut and scale the trees
cascade_defect <- merchandise(tree_id = trees$tree_id,
                              dbh = trees$dbh,
                              ht = trees$ht,
                              spcd = trees$spcd,
                              products = specifications,
                              age = trees$age,
                              defects = example_defects_pnw,
                              strategy = 'cascade')
```

```{r}
## Cut and scale the trees
optimized_defect <- merchandise(tree_id = trees$tree_id,
                                dbh = trees$dbh,
                                ht = trees$ht,
                                spcd = trees$spcd,
                                products = specifications,
                                age = trees$age,
                                defects = example_defects_pnw,
                                strategy = 'optimize')
```

```{r, echo = FALSE}
bind_rows(cascade = cascade_defect$logs, optimized = optimized_defect$logs, .id = 'strategy') %>%
  filter(tree_id %in% c(7, 14, 28)) %>%
  select(strategy, tree_id, log, product, start_height, length, value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The left drawings show cascade cuts. The right drawings show optimized cuts on the same stems.

```{r, echo = FALSE, fig.width = 16, fig.height = 9}
par(mfrow = c(3, 2), mar = c(4, 4, 1, 1))
for (id in c(7, 14, 28)) {
  plot(cascade_defect, tree_id = id, main = '', ylab = 'Stem radius (inches)')
  plot(optimized_defect, tree_id = id, main = '', ylab = 'Stem radius (inches)')
}
```

```{r}
## Sum cascade log values by tree
cascade_values_defect <- cascade_defect$logs %>%
  group_by(tree_id) %>%
  summarize(cascade_dollars = sum(value))

## Sum optimized log values by tree
optimized_values_defect <- optimized_defect$logs %>%
  group_by(tree_id) %>%
  summarize(optimized_dollars = sum(value))

## Calculate each tree's value gain
gain_defect <- trees %>%
  select(tree_id = tree_id) %>%
  left_join(y = cascade_values_defect,
            by = 'tree_id') %>%
  left_join(y = optimized_values_defect,
            by = 'tree_id') %>%
  mutate(cascade_dollars = coalesce(cascade_dollars, 0),
         optimized_dollars = coalesce(optimized_dollars, 0),
         gain_dollars = optimized_dollars - cascade_dollars)
```

```{r, echo = FALSE}
gain_defect %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Sum values and gains across the list
total_defect <- gain_defect %>%
  summarize(cascade_dollars = sum(cascade_dollars),
            optimized_dollars = sum(optimized_dollars),
            gain_dollars = sum(gain_dollars))
```

```{r, echo = FALSE}
total_defect %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Another price list

The alternative list raises the illustrative pulp price. A second `merchandise()` call recuts the same trees and defects under the new list.

```{r}
## Define a second complete price list
higher_pulp <- specifications %>%
  mutate(price = replace(x = price,
                         list = product == 'pulp',
                         values = 60))

## Cut and scale the trees again under the higher pulp price
optimized_higher <- merchandise(tree_id = trees$tree_id,
                                dbh = trees$dbh,
                                ht = trees$ht,
                                spcd = trees$spcd,
                                products = higher_pulp,
                                age = trees$age,
                                defects = example_defects_pnw,
                                strategy = 'optimize')

## Sum value across each price list
price_totals <- bind_rows(base = optimized_defect$logs,
                          higher_pulp = optimized_higher$logs,
                          .id = 'price_list') %>%
  group_by(price_list) %>%
  summarize(value_dollars = sum(value))
```

```{r, echo = FALSE}
price_totals %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The bars compare total log value after recutting under each price list.

```{r, echo = FALSE}
barplot(price_totals$value_dollars, names.arg = price_totals$price_list,
        col = '#0072B2', border = NA, xlab = 'Price list', ylab = 'Value (dollars)')
```

## Computation

The state records the cut position and counts for products with log limits. Counts restart at every cull and restriction boundary.
Each transition tries an eligible product and length. The objective is total log value.
Equal values favor fewer logs, then longer logs lower on the stem. Remaining ties use lower physical ends, then product names in byte order.
Values within the documented rounding tolerance are treated as ties.

Without log limits, work grows with positions times products times candidate lengths. Storing chosen paths also costs time and memory proportional to their log counts.
Log limits multiply the possible states by their count combinations. Each capped count ranges from zero through its limit, and only reached states are stored.
This remains exponential in the number of independently capped products. Trees run through the existing parallel engine.

The benchmark reports its machine and workload before timing each strategy.

```text
CPU: AMD Ryzen 9 3900X 12-Core Processor
Threads: 8
R: R version 4.3.3 (2024-02-29)
Workload: synthetic Douglas-fir, dbh 18 to 22 inches, height 76 to 84 feet.
Profiles repeat in a 45-tree cycle. Public caching is included.
Each timed call processes up to 1000 trees. No max_logs limits or defects.
```

| Products | Strategy | Stems | Threads | Seconds | Stems per second | Million stems, extrapolated seconds |
| ---: | :--- | ---: | ---: | ---: | ---: | ---: |
| 3 | cascade | 100000 | 8 | 337.284 | 296.486 | 3372.84 |
| 3 | optimize | 100000 | 8 | 315.510 | 316.947 | 3155.10 |
| 10 | cascade | 100000 | 8 | 290.461 | 344.280 | 2904.61 |
| 10 | optimize | 100000 | 8 | 336.459 | 297.213 | 3364.59 |

Smaller runs with distinct diameters expose the cost without repeated profiles. These runs also use eight threads.

| Products | Strategy | Distinct stems | Seconds | Stems per second | Million stems, extrapolated seconds |
| ---: | :--- | ---: | ---: | ---: | ---: |
| 3 | cascade | 1000 | 25.399 | 39.372 | 25399.00 |
| 3 | optimize | 1000 | 24.519 | 40.785 | 24519.00 |
| 10 | cascade | 1000 | 24.995 | 40.008 | 24995.00 |
| 10 | optimize | 1000 | 25.440 | 39.308 | 25440.00 |

Other validation tasks ran during these measurements.
The million-stem figures extrapolate these measured runs. They do not measure a million distinct stems or workloads with many independently capped products.
