---
title: 'Source agreement'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Source agreement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

The packaged agreement table records comparisons against the pinned National Volume Estimator Library build.
It reports precision, compatibility mode, compared rows, tolerance, maximum relative difference, and named exclusions.

`DIB` denotes diameter inside bark, and `DOB` denotes diameter outside bark.

```{r}
## Report the pinned source revision
nvel_source_revision()
```

```{r, echo = FALSE}
agreement <- read.csv(system.file('extdata',
                                  'oracle-agreement.csv',
                                  package = 'merchandiser'))
agreement %>%
  mutate(max_relative_difference = format(max_relative_difference, scientific = TRUE, digits = 12),
         tolerance = format(tolerance, scientific = TRUE, digits = 12)) %>%
  knitr::kable(row.names = FALSE)
```

Fixture tests enforce each tolerance and exclusion count. The full source fixtures live outside the package.
[Compatibility modes](methods-taper-equations.html) select package or source behavior for the documented differences.
