---
title: 'Pacific Northwest tree list'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pacific Northwest tree list}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product order in `products()` sets cutting priority. `merchandise()` produces product yield and residual sections from the complete Pacific Northwest tree list.

## Products and defects

The products reproduce `data-raw/example-products.R`. The trees and mapped defects are synthetic package examples.

```{r}
## Define export specifications
export <- product(product = 'export',  ## product name, unitless
                  spcd = c(202, 263),  ## species codes, unitless
                  min_dbh = 12,  ## inches
                  min_length = 32,  ## feet
                  max_length = 40,  ## feet
                  trim = 1,  ## feet
                  min_sed = 12,  ## inches
                  inside_bark = TRUE,  ## logical, unitless
                  volume_unit = 'scribner',  ## board feet
                  split_scale = FALSE)  ## logical, unitless

## Define domestic saw specifications
domestic_saw <- product(product = 'domestic_saw',  ## product name, unitless
                        spcd = c(202, 263),  ## species codes, unitless
                        min_dbh = 10,  ## inches
                        min_length = 16,  ## feet
                        max_length = 40,  ## feet
                        trim = 1,  ## feet
                        min_sed = 6,  ## inches
                        inside_bark = TRUE,  ## logical, unitless
                        volume_unit = 'scribner',  ## board feet
                        split_scale = FALSE)  ## logical, unitless

## Define pulp specifications
pulp <- product(product = 'pulp',  ## product name, unitless
                spcd = c(202, 263),  ## species codes, unitless
                min_dbh = 5,  ## inches
                min_length = 8,  ## feet
                max_length = 40,  ## feet
                trim = 0.5,  ## feet
                min_sed = 3,  ## inches
                inside_bark = TRUE,  ## logical, unitless
                volume_unit = 'green_ton')  ## green short tons

## Combine products in cutting priority order
specifications <- products(export = export,
                           domestic_saw = domestic_saw,
                           pulp = pulp)

## Retain the complete example list
trees <- example_trees_pnw

## Cut and scale the trees
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = specifications,
                      defects = example_defects_pnw,
                      age = trees$age)

## Join log rows to tree measurements
logs <- result$logs %>%
  left_join(y = trees,
            by = c('tree_id' = 'tree_id'))
```

## Product yield

Scale is summed across the listed trees. No expansion factors are supplied with this list.

```{r, echo = FALSE}
logs %>%
  filter(volume_unit == 'scribner') %>%
  group_by(product) %>%
  summarize(`logs (count)` = n(), `scale (Scribner board feet)` = sum(scale)) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
logs %>%
  filter(volume_unit == 'green_ton') %>%
  group_by(product) %>%
  summarize(`logs (count)` = n(), `scale (green short tons)` = sum(scale)) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The stacked bars show which diameter classes contribute each product. Separate panels keep the scale units distinct.

```{r, echo = FALSE, fig.width = 9, fig.height = 7}
product_names <- specifications$product
product_colors <- setNames(vapply(product_names, function(name) {
  code <- utf8ToInt(enc2utf8(name))
  grDevices::hcl(sum(code * seq_along(code)) %% 360, c = 65, l = 55)
}, character(1)), product_names)
yield <- logs %>%
  mutate(diameter_class = floor(dbh / 4) * 4) %>%
  group_by(product, volume_unit, diameter_class) %>%
  summarize(scale = sum(scale), .groups = 'drop')
yield %>%
  filter(volume_unit == 'scribner') %>%
  transmute(product, `diameter class lower bound (inches)` = diameter_class,
            `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
yield %>%
  filter(volume_unit == 'green_ton') %>%
  transmute(product, `diameter class lower bound (inches)` = diameter_class,
            `scale (green short tons)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE, fig.width = 9, fig.height = 7}
par(mfrow = c(2, 1), mar = c(4, 5, 1, 1))
for (unit in c('scribner', 'green_ton')) {
  z <- yield %>% filter(volume_unit == unit)
  amounts <- xtabs(scale ~ product + diameter_class, data = z)
  label <- if (unit == 'scribner') 'Scale (Scribner board feet)' else 'Scale (green short tons)'
  barplot(amounts, col = product_colors[rownames(amounts)], border = NA,
          xlab = 'Diameter class lower bound (inches)', ylab = label)
  legend('topleft', legend = rownames(amounts), fill = product_colors[rownames(amounts)],
         bty = 'n', cex = 0.8)
}
```

## Selected stems

<details>
<summary>Drawing measurements</summary>

```{r, echo = FALSE}
drawing_trees <- trees %>%
  filter(tree_id %in% c(7, 14, 28))
drawing_rows <- match(drawing_trees$tree_id, result$call$tree_id)
drawing_models <- result$call$model[drawing_rows]
drawing_profiles <- stem_profile(
  tree_id = drawing_trees$tree_id,
  dbh = drawing_trees$dbh,
  ht = drawing_trees$ht,
  spcd = drawing_trees$spcd,
  model = drawing_models,
  step = 0.5,
  from = 0
)
drawing_profiles %>%
  transmute(tree_id, `height (feet)` = h, `radius inside bark (inches)` = dib / 2) %>%
  knitr::kable(digits = 4, row.names = FALSE)
result$residuals %>%
  filter(tree_id %in% drawing_trees$tree_id) %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height) %>%
  knitr::kable(digits = 3, row.names = FALSE)
result$logs %>%
  filter(tree_id %in% drawing_trees$tree_id, volume_unit == 'scribner') %>%
  transmute(tree_id, log, product,
            `start height (feet)` = start_height, `end height (feet)` = end_height,
            `length (feet)` = length, `small end diameter (inches)` = sed,
            `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
result$logs %>%
  filter(tree_id %in% drawing_trees$tree_id, volume_unit == 'green_ton') %>%
  transmute(tree_id, log, product,
            `start height (feet)` = start_height, `end height (feet)` = end_height,
            `length (feet)` = length, `small end diameter (inches)` = sed,
            `scale (green short tons)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
example_defects_pnw %>%
  filter(tree_id %in% drawing_trees$tree_id) %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height,
         `percent (percent)` = percent) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

</details>

The stacked drawings locate products and mapped defects on selected trees from the run.

```{r, echo = FALSE, fig.width = 10, fig.height = 9}
par(mfrow = c(3, 1), mar = c(4, 4, 1, 1))
plot(result, tree_id = c(7, 14, 28), ylab = 'Stem radius (inches)')
```

## Checks

```{r, echo = FALSE}
result$status %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
result$residuals %>%
  group_by(cause) %>%
  summarize(`sections (count)` = n(), `total length (feet)` = sum(end_height - start_height)) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

[Completing heights](fill-missing-heights.html) reproduces the predicted heights already stored in this list.
