---
title: 'Scaling rules'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product scaling fields select the quantity and the measurements used to calculate it.
The returned log columns retain the bark basis, scaling diameter, and scaling length.

```{r}
## Select example trees
trees <- example_trees %>%
  slice_head(n = 3)
```

## `volume_unit`

Choices are `scribner`, `international`, `doyle`, `cubic`, `green_ton`, and `cord`.
Board foot rules use inside bark diameters. Green ton scale includes wood and attached bark.

## `inside_bark`

This field selects the bark basis for diameter limits. Cubic and cord scales use that same basis.

## `split_scale`

This field selects shorter scaling sections for Scribner. It does not add physical cuts or affect other scale units.

## `round`

Choices are `default`, `down`, `nearest`, and `none`. They control diameter preprocessing before the rule's own rounding.
`none` therefore does not disable rounding inside a source rule.

## `length_round`

Nominal length rounds down to this scaling increment. Source board foot rules accept whole or even feet.
Cubic, green ton, and cord scales use the nominal body and report scaling length for the measurement record.

```{r}
## Define saw specifications
saw <- product(product = 'domestic_saw',  ## product name, unitless
               min_length = 31.5,  ## feet
               max_length = 31.5,  ## feet
               min_sed = 6,  ## inches
               volume_unit = 'scribner',  ## Scribner board feet
               inside_bark = TRUE,  ## logical, unitless
               split_scale = TRUE,  ## logical, unitless
               round = 'down',  ## rounding choice, unitless
               length_round = 2)  ## feet

## Cut and scale the trees
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = saw)
```

```{r, echo = FALSE}
result$logs %>%
  transmute(tree_id, log, `length (feet)` = length,
            `scaling_length (feet)` = scaling_length, `sed (inches)` = sed,
            `scaling_diameter (inches)` = scaling_diameter, inside_bark,
            `scale (Scribner board feet)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```
