---
title: 'Southern stands'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Southern stands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Product order in `products()` sets cutting priority. `merchandise()` produces green short ton yield by product and stand from the southern tree list.

## Products

The products match `data-raw/example-products.R`.

```{r}
## Define sawtimber specifications
sawtimber <- product(product = 'sawtimber',  ## product name, unitless
                     spcd = 131,  ## species code, unitless
                     min_dbh = 12,  ## inches
                     min_length = 16,  ## feet
                     max_length = 40,  ## feet
                     trim = 0.5,  ## feet
                     min_sed = 8,  ## inches
                     inside_bark = FALSE,  ## logical, unitless
                     max_logs = 1,  ## logs per segment
                     volume_unit = 'green_ton')  ## green short tons

## Define chip n saw specifications
chip_n_saw <- product(product = 'chip_n_saw',  ## product name, unitless
                      spcd = 131,  ## species code, unitless
                      min_dbh = 8,  ## inches
                      max_dbh = 12,  ## inches
                      min_length = 16,  ## feet
                      max_length = 40,  ## feet
                      trim = 0.5,  ## feet
                      min_sed = 6,  ## inches
                      inside_bark = FALSE,  ## logical, unitless
                      max_logs = 1,  ## logs per segment
                      volume_unit = 'green_ton')  ## green short tons

## Define pulpwood specifications
pulpwood <- product(product = 'pulpwood',  ## product name, unitless
                    spcd = 131,  ## species code, unitless
                    min_dbh = 5,  ## inches
                    max_dbh = 8,  ## inches
                    min_length = 8,  ## feet
                    max_length = 40,  ## feet
                    trim = 0.5,  ## feet
                    min_sed = 3,  ## inches
                    inside_bark = FALSE,  ## logical, unitless
                    max_logs = 1,  ## logs per segment
                    volume_unit = 'green_ton')  ## green short tons

## Combine products in cutting priority order
specifications <- products(sawtimber = sawtimber,
                           chip_n_saw = chip_n_saw,
                           pulpwood = pulpwood)

## Retain the complete southern list
trees <- example_trees_south
```

## Stopper records

```{r}
## Convert the stopper columns
records <- defects_from_stoppers(
  tree_id = trees$tree_id,
  ht = trees$ht,
  topwood_product = "pulpwood",
  saw_stop = trees$saw_stop,
  pulp_stop = trees$pulp_stop,
  jump_butt = trees$jump_butt
)
```

```{r, echo = FALSE}
records %>%
  slice_head(n = 6) %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height,
         `percent (%)` = percent) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Cut and scale the trees
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = specifications,
                      age = trees$age,
                      defects = records)

## Join logs to stand and diameter
logs <- result$logs %>%
  left_join(y = trees,
            by = c('tree_id' = 'tree_id'))
```

## Product yield

The bars show scale by diameter class within each stand. Totals represent the listed trees, without expansion.

```{r, echo = FALSE, fig.width = 9, fig.height = 7}
product_names <- specifications$product
product_colors <- setNames(vapply(product_names, function(name) {
  code <- utf8ToInt(enc2utf8(name))
  grDevices::hcl(sum(code * seq_along(code)) %% 360, c = 65, l = 55)
}, character(1)), product_names)
yield <- logs %>%
  mutate(diameter_class = floor(dbh / 2) * 2) %>%
  group_by(stand, age, product, diameter_class) %>%
  summarize(scale = sum(scale), .groups = 'drop')
yield %>%
  rename(`age (years)` = age,
         `diameter class lower bound (inches)` = diameter_class,
         `scale (green short tons)` = scale) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE, fig.width = 9, fig.height = 7}
par(mfrow = c(2, 2), mar = c(4, 4, 1, 1))
for (stand_id in unique(yield$stand)) {
  z <- yield %>% filter(stand == stand_id)
  amounts <- xtabs(scale ~ product + diameter_class, data = z)
  barplot(amounts, col = product_colors[rownames(amounts)], border = NA,
          xlab = 'Diameter class lower bound (inches)', ylab = 'Scale (green short tons)')
  legend('topleft', legend = c(paste('Age', unique(z$age), 'years'), rownames(amounts)),
         fill = c(NA, product_colors[rownames(amounts)]), bty = 'n', cex = 0.75)
}
```

```{r, echo = FALSE}
logs %>%
  group_by(stand, age, product) %>%
  summarize(`scale (green short tons)` = sum(scale), .groups = 'drop') %>%
  rename(`age (years)` = age) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Age limits

The source products have no age limits. Explicit age limits further restrict product eligibility.

```{r}
## Define sawtimber specifications
sawtimber <- product(product = 'sawtimber',  ## product name, unitless
                     min_dbh = 12,  ## inches
                     min_age = 35,  ## years
                     min_length = 16,  ## feet
                     max_length = 40,  ## feet
                     trim = 0.5,  ## feet
                     min_sed = 8,  ## inches
                     inside_bark = FALSE,  ## logical, unitless
                     max_logs = 1,  ## logs per segment
                     volume_unit = 'green_ton')  ## green short tons

## Define chip n saw specifications
chip_n_saw <- product(product = 'chip_n_saw',  ## product name, unitless
                      min_dbh = 8,  ## inches
                      min_age = 20,  ## years
                      max_dbh = 12,  ## inches
                      min_length = 16,  ## feet
                      max_length = 40,  ## feet
                      trim = 0.5,  ## feet
                      min_sed = 6,  ## inches
                      inside_bark = FALSE,  ## logical, unitless
                      max_logs = 1,  ## logs per segment
                      volume_unit = 'green_ton')  ## green short tons

## Define pulpwood specifications
pulpwood <- product(product = 'pulpwood',  ## product name, unitless
                    min_dbh = 5,  ## inches
                    max_age = 20,  ## years
                    max_dbh = 8,  ## inches
                    min_length = 8,  ## feet
                    max_length = 40,  ## feet
                    trim = 0.5,  ## feet
                    min_sed = 3,  ## inches
                    inside_bark = FALSE,  ## logical, unitless
                    max_logs = 1,  ## logs per segment
                    volume_unit = 'green_ton')  ## green short tons

## Combine products in cutting priority order
specifications <- products(sawtimber = sawtimber,
                           chip_n_saw = chip_n_saw,
                           pulpwood = pulpwood)

## Cut and scale the trees
age_result <- merchandise(tree_id = trees$tree_id,
                          dbh = trees$dbh,
                          ht = trees$ht,
                          spcd = trees$spcd,
                          products = specifications,
                          age = trees$age,
                          defects = records)
```

```{r, echo = FALSE}
age_result$logs %>%
  left_join(y = trees,
            by = c('tree_id' = 'tree_id')) %>%
  group_by(stand, age, product) %>%
  summarize(`scale (green short tons)` = sum(scale), .groups = 'drop') %>%
  rename(`age (years)` = age) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Age limits change which products qualify. Diameter limits and stopper restrictions still apply.

## Status

Trees without an eligible product remain in the status table.

```{r, echo = FALSE}
age_result$status %>%
  count(name, name = 'trees (count)') %>%
  knitr::kable(digits = 3, row.names = FALSE)
```
