---
title: 'Stem model calculations'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stem model calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

Scalar height, diameter, volume, and biomass helpers return results aligned with their input vectors.
`stem_profile()` expands trees into rows sorted by identifier and height.
Only length-one inputs recycle. Other nonempty lengths must match.

## Heights

```{r}
## Fit the retained inventory heights
fit <- fit_height(dbh = example_trees_pnw$dbh,
                  ht = example_trees_pnw$ht_observed,
                  spcd = example_trees_pnw$spcd)
```

```{r}
## Complete missing heights with the fitted model
completed <- example_trees_pnw %>%
  mutate(ht = complete_heights(dbh = dbh,
                               ht = ht_observed,
                               spcd = spcd,
                               fit = fit))
```

```{r, echo = FALSE}
completed %>%
  select(tree, ht_observed, ht) %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Stem quantities

`dib()` calculates inside-bark diameter. `height_at_dib()` returns the highest matching height.
`stem_volume()` evaluates a bounded interval using the model's volume capability.

```{r}
## Add diameter, height, and volume calculations
dimensions <- example_trees %>%
  transmute(tree,
            diameter = dib(dbh = dbh, ht = ht, h = 20, model = model),
            height = height_at_dib(dbh = dbh, ht = ht, dib = 6, model = model),
            cubic = stem_volume(dbh = dbh, ht = ht, model = model))
```

```{r, echo = FALSE}
dimensions %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Imperial diameters, heights, and volumes use inches, feet, and cubic feet.
Metric calls require centimeters and meters and return cubic meters.
Outside-bark calculations require a model capability or an explicit constant bark ratio.

## Weight

```{r}
## Convert the modeled volumes to green wood and bark weight
weight <- green_weight(volume = dimensions$cubic,
                       spcd = example_trees$spcd,
                       status = TRUE)
```

```{r, echo = FALSE}
weight %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Green weight uses species properties and supplied cubic volume.
The national biomass equations use tree dimensions and ecological divisions to calculate fitted component masses.
[biomass()](biomass-carbon.html) returns dry mass, green mass, and carbon by component.

`status = TRUE` pairs quantities with diagnoses.
Unsupported quantities remain missing. Model checks verify available operations and species scope, not predictive accuracy.
