---
title: 'Stem model calculations'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stem model calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Stem measurement results align with input rows and can be joined to tree lists.
`stem_profile()` expands each tree into ordered height measurements.

## Input lengths

Only scalar inputs recycle. Other nonempty input lengths must match.

## Heights

```{r}
## Select measured rows
measured <- example_trees_pnw %>%
  filter(ht_status == 'measured')

## Fit measured heights
fit <- fit_height(dbh = measured$dbh,
                  ht = measured$ht,
                  spcd = measured$spcd,
                  group = measured$plot)

## Blank predicted rows before completion
trees <- example_trees_pnw %>%
  mutate(ht = if_else(condition = ht_status == 'predicted',
                      true = NA_real_,
                      false = ht))

## Complete the missing heights
completed <- trees %>%
  mutate(ht = complete_heights(dbh = dbh,
                               ht = ht,
                               spcd = spcd,
                               group = plot,
                               fit = fit))
```

```{r, echo = FALSE}
completed %>%
  slice_head(n = 5) %>%
  transmute(tree_id = tree_id, ht_status, `height (feet)` = ht) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

## Stem quantities

Each measurement function returns `value` and `status`. Outside bark calculations need a model capability or a bark ratio.

```{r}
## Calculate dib
dib_result <- dib(dbh = example_trees$dbh,
                  ht = example_trees$ht,
                  spcd = example_trees$spcd,
                  model = example_trees$model,
                  h = 20)

## Calculate height at dib
height_at_dib_result <- height_at_dib(dbh = example_trees$dbh,
                                      ht = example_trees$ht,
                                      spcd = example_trees$spcd,
                                      model = example_trees$model,
                                      dib = 6)

## Calculate stem volume
stem_volume_result <- stem_volume(dbh = example_trees$dbh,
                                  ht = example_trees$ht,
                                  spcd = example_trees$spcd,
                                  model = example_trees$model)
```

```{r, echo = FALSE}
example_trees %>%
  transmute(tree_id = tree_id, `diameter inside bark (inches)` = dib_result$value,
            `height at target diameter (feet)` = height_at_dib_result$value,
            `volume (cubic feet)` = stem_volume_result$value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Diameters use inches, heights use feet, and volumes use cubic feet. Inverse functions retain the highest matching height.

## Weight

```{r}
## Convert modeled stem volume to weight
weight <- green_weight(volume = stem_volume_result$value,
                       spcd = example_trees$spcd)
```

```{r, echo = FALSE}
weight %>%
  rename(`green weight (short tons)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

[Green weight](volumes-scaling.html) uses supplied cubic volume and species properties. [Biomass](biomass-carbon.html) estimates dry mass and carbon from tree dimensions.

Unsupported quantities remain missing with status codes. Model checks establish available operations and species scope, not predictive accuracy.
