---
title: 'Taper and stem profiles'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taper and stem profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

Stem functions accept measurement vectors and equation identifiers.
Default diameters use inches, heights use feet, and volumes use cubic feet.
`units = 'metric'` requires centimeters and meters and returns cubic meters.

`step` uses diameter units: inches for imperial calls and centimeters for metric calls.

## Materialize a profile

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Calculate a stem profile
profile <- stem_profile(dbh = tree$dbh,
                        ht = tree$ht,
                        model = tree$model,
                        id = tree$tree,
                        step = 10,
                        status = TRUE)
```

```{r, echo = FALSE}
profile %>%
  slice_head(n = 6) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Draw the modeled profile
plot(profile)
```

The profile contains inside- and outside-bark diameters and cumulative volumes from its lower bound.
`status` is always present. `status = TRUE` suppresses row warnings. Unsupported outside-bark quantities remain missing.

## Diameters and heights

```{r}
## Estimate diameter on both bark bases
diameters <- example_trees %>%
  transmute(tree,
            inside = dib(dbh = dbh, ht = ht, h = 20, model = model),
            outside = dob(dbh = dbh, ht = ht, h = 20, model = model))
```

```{r, echo = FALSE}
diameters %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Find heights at the specified diameters
heights <- example_trees %>%
  transmute(tree,
            inside = height_at_dib(dbh = dbh, ht = ht, dib = 6, model = model),
            outside = height_at_dob(dbh = dbh, ht = ht, dob = 6, model = model))
```

```{r, echo = FALSE}
heights %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

The inverse functions return the highest crossing. Additional crossings produce a diagnosis with the retained height.

## Equations and registry

```{r}
## List equations for the example species
models <- taper_models(species = tree$spcd)
```

```{r, echo = FALSE}
models %>%
  select(id, family, has_dob, has_inverse, has_integral) %>%
  slice_head(n = 6) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Inspect the selected equation and its capabilities
get_taper_model(id = tree$model)

## Check the equation's available operations
capabilities <- model_capabilities(id = tree$model)
```

```{r, echo = FALSE}
capabilities %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Check availability and the species assignment
has_taper_model(id = tree$model)

## Validate the equation mapping
check_taper_models(id = tree$model, spcd = tree$spcd)
```

[Source equations](methods-taper-equations.html) support package and source compatibility modes.
[fit_taper()](methods-fitting-taper.html) estimates coefficients for session registration.
