---
title: 'Taper and stem profiles'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taper and stem profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

Stem functions return diameters, heights, and cumulative volumes for checking the modeled stem.
Inputs use inches and feet, and volumes use cubic feet.

```{r}
## Select a tree for the profile
tree <- example_trees %>%
  filter(tree_id == 5)

## Measure the stem profile
profile <- stem_profile(
  tree_id = tree$tree_id,
  dbh = tree$dbh,
  ht = tree$ht,
  spcd = tree$spcd,
  model = tree$model,
  step = 1
)
```

`step` is spacing in feet along the stem. Cumulative volume starts at the lower bound.

<details>
<summary>Stem profile measurements</summary>

```{r, echo = FALSE}
profile %>%
  rename(`h (feet)` = h, `dib (inches)` = dib, `dob (inches)` = dob,
         `cum_volume_ib (cubic feet)` = cum_volume_ib,
         `cum_volume_ob (cubic feet)` = cum_volume_ob) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

</details>

The solid and dashed lines show inside and outside bark diameters.

```{r, echo = FALSE}
plot(profile, xlim = range(c(profile$dib, profile$dob), na.rm = TRUE))
```

## Diameters and heights

```{r}
## Calculate dib
dib(dbh = tree$dbh,
    ht = tree$ht,
    h = 20,
    spcd = tree$spcd,
    model = tree$model) %>%
  rename(`diameter inside bark (inches)` = value)

## Calculate dob
dob(dbh = tree$dbh,
    ht = tree$ht,
    h = 20,
    spcd = tree$spcd,
    model = tree$model) %>%
  rename(`diameter outside bark (inches)` = value)

## Calculate height at dib
height_at_dib(dbh = tree$dbh,
              ht = tree$ht,
              dib = 6,
              spcd = tree$spcd,
              model = tree$model) %>%
  rename(`height (feet)` = value)

## Calculate height at dob
height_at_dob(dbh = tree$dbh,
              ht = tree$ht,
              dob = 6,
              spcd = tree$spcd,
              model = tree$model) %>%
  rename(`height (feet)` = value)
```

Diameter results use inches. Inverse results use feet and retain the highest crossing when several exist.
Status identifies unavailable measurements and multiple crossings.

## Taper registry

`has_inverse` records a model-provided inverse. Numeric inversion remains available for the selected model when this flag is false.

```{r}
## List models for the selected species
models <- taper_models(spcd = tree$spcd)
```

```{r, echo = FALSE}
models %>%
  select(id, form, has_dob, has_inverse, has_integral) %>%
  slice_head(n = 6) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Inspect the selected model
get_taper_model(model = tree$model)$id

## Check model capabilities
taper_models() %>%
  filter(id == tree$model) %>%
  select(id, has_dob, has_inverse, has_integral)

## Check availability and species scope
has_taper_model(model = tree$model)

## Check the assignment
check_taper_models(model = tree$model, spcd = tree$spcd)
```

[Fitting taper equations](methods-fitting-taper.html) creates local models for the same stem functions.
