---
title: 'Volumes and green weight'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Volumes and green weight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 120)
library(merchandiser)
library(dplyr)
```

`stem_volume()` measures bounded stem sections and volume above the stump without product specifications.
`green_weight()` converts cubic volume to short tons for merchandising.

```{r}
## Select example trees
trees <- example_trees %>%
  slice_head(n = 3)

## Measure a section inside bark
section <- stem_volume(
  dbh = trees$dbh,
  ht = trees$ht,
  spcd = trees$spcd,
  from = 1,
  to = 32
)
```

```{r, echo = FALSE}
section %>%
  rename(`volume (cubic feet)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Bounds may be heights, inside or outside bark diameters, the stump, or the tip. `inside_bark` selects the volume basis.

```{r}
## Measure each tree above the stump
volume <- stem_volume(
  dbh = trees$dbh,
  ht = trees$ht,
  spcd = trees$spcd,
  stump_ht = 1,
  to_dib = 6
)
```

```{r, echo = FALSE}
volume %>%
  rename(`volume (cubic feet)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Omitting `to_dib` measures to the tip. The top diameter stays inside bark when `inside_bark = FALSE`.

## Green weight

The default converts inside bark volume to green wood and attached bark weight using species properties.

```{r}
## Convert stem volume to green short tons
weight <- green_weight(volume = volume$value,
                       spcd = trees$spcd)
```

```{r, echo = FALSE}
weight %>%
  rename(`green weight (short tons)` = value) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`component`, `moisture`, and the physical property overrides control the conversion. `inside_bark = FALSE` accepts volume including bark.

## Trim

A log occupies its nominal length plus trim, is scaled on its nominal length, and leaves the trim wood as a residual.

```{r}
## Define saw specifications
saw <- product(product = 'domestic_saw',  ## product name, unitless
               min_length = 16,  ## feet
               max_length = 40,  ## feet
               trim = 1,  ## feet
               min_sed = 6,  ## inches
               volume_unit = 'scribner')  ## Scribner board feet

## Cut logs with trim
result <- merchandise(tree_id = trees$tree_id,
                      dbh = trees$dbh,
                      ht = trees$ht,
                      spcd = trees$spcd,
                      products = saw)
```

```{r, echo = FALSE}
result$logs %>%
  filter(volume_unit == 'scribner') %>%
  transmute(`tree_id` = tree_id,
            `log` = log,
            `product` = product,
            `start_height (feet)` = start_height,
            `end_height (feet)` = end_height,
            `length (feet)` = length,
            `scaling_length (feet)` = scaling_length,
            `sed (inches)` = sed,
            `led (inches)` = led,
            `scaling_diameter (inches)` = scaling_diameter,
            `inside_bark` = inside_bark,
            `scale (Scribner board feet)` = scale,
            `volume_unit` = volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r, echo = FALSE}
result$residuals %>%
  filter(cause == 'trim') %>%
  rename(`start_height (feet)` = start_height, `end_height (feet)` = end_height) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

[Scaling rules](scaling-rules.html) control how nominal lengths and diameters enter the selected scale.
