---
title: 'Volumes and scaling'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Volumes and scaling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = '#>',
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 4)
library(merchandiser)
library(dplyr)
```

`stem_volume()` calculates volume between stem bounds. `tree_cubic_volume()` returns the physical volume of selected logs.
The example currency is United States dollars (`USD`).
Product measurements are stored separately in `gross_scale` and `net_scale`.

## Stem and log volume

```{r}
## Calculate volume from stump to a specified inside-bark top
volumes <- example_trees %>%
  transmute(tree,
            stem_cubic = stem_volume(dbh = dbh,
                                     ht = ht,
                                     model = model,
                                     upper = 6,
                                     upper_type = 'dib'))
```

```{r, echo = FALSE}
volumes %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Calculate physical log volume under the example specifications
log_volumes <- example_trees %>%
  transmute(tree,
            log_cubic = tree_cubic_volume(dbh = dbh,
                                          ht = ht,
                                          species = species,
                                          model = model,
                                          products = example_products(name = 'pnw')))
```

```{r, echo = FALSE}
log_volumes %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`stem_volume()` accepts height, inside-bark diameter, outside-bark diameter, stump, or tip bounds.
Its default interval runs from the model's stump height to the tip.
Physical log volume includes trim. Product scaling uses nominal length and excludes trim.

## Measurement fields

`volume_unit` selects Scribner, International, Doyle, cubic volume, green tons, or cords.
`inside_bark` sets acceptance diameter basis. Cubic and cord measurements use that bark basis.

Board-foot measurements use inside bark. Green weight includes wood and attached bark.

`split_scale` selects whole-log or section scaling for Scribner. It creates measurement sections without additional physical cuts.
`round` selects default, downward, nearest-inch, or unrounded diameter preprocessing.
The underlying board-foot procedure retains its prescribed internal rounding.

```{r}
## Select a shipped tree
tree <- example_trees %>%
  filter(tree == 5)
```

```{r, echo = FALSE}
tree %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Load the example specifications
specifications <- example_products(name = 'douglas_fir')
```

```{r, echo = FALSE}
specifications %>%
  mutate(lengths = vapply(lengths, paste, character(1), collapse = ', ')) %>%
  select(product, lengths, min_length, max_length, min_sed, trim, volume_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Select logs from the tree
result <- merchandise(dbh = tree$dbh,
                      ht = tree$ht,
                      model = tree$model,
                      species = tree$species,
                      id = tree$tree,
                      products = specifications,
                      currency = 'USD',
                      status = TRUE,
                      report_also = c('smalian', 'huber', 'international', 'doyle'))
```

```{r, echo = FALSE}
result$logs %>%
  select(log, product, nominal_length, sed_ib, net_scale, scale_unit) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
## Select additional measurements on the existing cuts
measurements <- result$scales %>%
  filter(!is_product) %>%
  select(log, scale_rule, gross, net, unit, bark_basis)
```

```{r, echo = FALSE}
measurements %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

`report_also` measures existing cuts with additional procedures. It does not change product acceptance or bucking.

Smalian and Huber reports use the nominal log body. Named definitions support fixed units and additional rounding operators.
[Scaling rules](../reference/scaling_rules.html) map product fields to measurement procedures.

Green tons follow the call's units: short tons for imperial and metric tons for metric.
Cord measurement requires `cord_solid_fraction`. [green_weight()](biomass-carbon.html) converts cubic volume to weight.
